---
title: "Basic R Programming Practice"
author: "Tim Irwin"
date: "2026-04-16"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Basic R Programming Practice

This document is for practice based on the swirl R Programming lessons:

1. Basic Building Blocks  
2. Workspace and Files  
3. Sequences of Numbers  
4. Vectors  
5. Missing Values  
6. Subsetting Vectors  
7. Matrices and Data Frames  

Write your answers **inside the R chunks** below.

You may want to start with:

```{r}
rm(list = ls())
setwd("/Users/timothyirwin/Desktop/Coursera/Foundations/data science/Practice")
```

---

# 1. Basic Building Blocks

## Exercise 1.1
Assign the value `12` to an object called `x`.

```{r}
x <- 12
```

## Exercise 1.2
Assign the value `3.5` to an object called `y`.

```{r}
y <- 3.5
```

## Exercise 1.3
Add `x` and `y`.

```{r}
x + y
```

## Exercise 1.4
Create an object called `name` that stores your first name as a character string.

```{r}
name <- "Tim"
```

## Exercise 1.5
Check the class of `x`, `y`, and `name`.

```{r}
class(x)
class(y)
class(name)
```

---

# 2. Workspace and Files

## Exercise 2.1
Display your current working directory.

```{r}
getwd()
```

## Exercise 2.2
Assign your current working directory to an object called `old_dir`.

```{r}
old_dir <- getwd()
```

## Exercise 2.3
Create a new directory called `practice_folder`.

```{r}
dir.create("practice_folder")
```

## Exercise 2.4
List the files in your current working directory.

```{r}
ls()
```

## Exercise 2.5
Delete `practice_folder` when you are done.

```{r}
unlink("practice_folder", recursive = TRUE)
```

---

# 3. Sequences of Numbers

## Exercise 3.1
Create a sequence from 1 to 10.

```{r}
1:10
```

## Exercise 3.2
Create a sequence from 0 to 20 by 2.

```{r}
seq(0, 20, by = 2)
```

## Exercise 3.3
Create a sequence of 8 numbers from 5 to 40.

```{r}
seq(5, 40, length.out = 8)
```

## Exercise 3.4
Suppose:

```{r}
my_seq <- c(4, 8, 12, 16, 20)
my_seq
```

Create a sequence from 1 to the length of `my_seq`.

```{r}
seq_along(my_seq)
```

---

# 4. Vectors

## Exercise 4.1
Create a numeric vector called `num_vect` containing `0.5`, `55`, `-10`, and `6`.

```{r}
num_vect <- c(0.5, 55, -10, 6)
num_vect
```

## Exercise 4.2
Create a logical vector called `tf` that tests whether `num_vect` is less than 1.

```{r}
tf <- num_vect < 1
tf
```

## Exercise 4.3
Create a character vector called `days` with the values `"Mon"`, `"Tue"`, `"Wed"`, `"Thu"`, `"Fri"`.

```{r}
days <- c("Mon", "Tue", "Wed", "Thu", "Fri")
days
```

## Exercise 4.4
Combine `num_vect` with itself and the number `999` in the middle.

```{r}
c(num_vect, 999, num_vect)
```

## Exercise 4.5
Multiply `num_vect` by 2 and then add 100.

```{r}
(num_vect * 2) + 100
```

---

# 5. Missing Values

## Exercise 5.1
Create a vector called `trial` with values `1`, `5`, `NA`, `10`, and `NA`.

```{r}
trial <- c(1, 5, NA, 10, NA)
trial
```

## Exercise 5.2
Use `is.na()` to check which values in `trial` are missing.

```{r}
is.na(trial)
```

## Exercise 5.3
Count how many missing values are in `trial`.

```{r}
sum(is.na(trial))
```

## Exercise 5.4
Find the mean of `trial`, excluding missing values.

```{r}
mean(trial, na.rm = TRUE)
```

---

# 6. Subsetting Vectors

## Exercise 6.1
Create a vector called `scores` with the values `88`, `72`, `95`, `64`, `90`.

```{r}
scores <- c(88, 72, 95, 64, 90)
scores
```

## Exercise 6.2
Extract the third element of `scores`.

```{r}
scores[3]
```

## Exercise 6.3
Extract the first three elements of `scores`.

```{r}
scores[1:3]
```

## Exercise 6.4
Extract all elements of `scores` greater than 80.

```{r}
scores[scores >80]
```

## Exercise 6.5
Replace all elements of `scores` less than 70 with `70`.

```{r}
scores[scores < 70] <- 70
scores
```

---

# 7. Matrices and Data Frames

## Exercise 7.1
Create two vectors:

```{r}
a <- c(1, 3, 5)
b <- c(2, 4, 6)
```

Use `cbind()` to combine them into a matrix.

```{r}
cbind(a, b)
```

## Exercise 7.2
Use `rbind()` to combine `a` and `b` into a matrix.

```{r}
rbind(a, b)
```

## Exercise 7.3
Create a data frame called `student_data` with these columns:
- `name`: `"Ana"`, `"Ben"`, `"Cara"`
- `score`: `91`, `85`, `88`
- `passed`: `TRUE`, `TRUE`, `TRUE`

```{r}
student_data <- data.frame(
  name = c("Ana", "Ben", "Cara"),
  score = c(91, 85, 88),
  passed = c(TRUE, TRUE, TRUE)
)
student_data
```

## Exercise 7.4
Display the names of the columns in `student_data`.

```{r}
names(student_data)
```

## Exercise 7.5
Extract the `score` column from `student_data`.

```{r}
student_data[ , 2]
```

## Exercise 7.6
Extract the second row of `student_data`.

```{r}
student_data[2, ]
```

---

# Mini Challenge

## Challenge 1
Create a vector from 1 to 20. Extract only the even numbers.

```{r}
d <- 1:20
d[seq(2, 20, by = 2)]
```

## Challenge 2
Create a vector with some missing values and calculate its mean without the missing values.

```{r}
c <- c(3, 7, NA, 10, 2, NA, 14, NA)
mean(c, na.rm = TRUE)
```

## Challenge 3
Create a 3 x 2 matrix using the numbers 1 through 6.

```{r}
matrix(1:6, nrow = 3, ncol = 2)
```
