---
title: "R Programming Drill: Logical Masks vs. Values"
author: "Tim Irwin"
date: "2026-05-20"
output:
  pdf_document: default
  html_document: default
---

# R Programming Drill: Logical Masks vs. Values

This is a focused drill — not a broad practice set. Every exercise targets one of two specific habits:

1. **Logical mask vs. vector of values** — knowing when you have a TRUE/FALSE mask versus actual data, and what each is good for.
2. **Store-as-you-go** — capturing intermediate results in variables instead of letting expressions evaporate.

The exercises are short and deliberately repetitive. That repetition is the point — you're building a reflex, not learning new material. Do them in order; they escalate slightly.

```{r}
rm(list = ls())
```

## The one idea behind all of this

**Comparison operators** (`>`, `<`, `>=`, `<=`, `==`, `!=`) turn values into TRUE/FALSE.
They answer a yes/no question about each element. The result is called a **logical mask**.

**Logical operators** (`&`, `|`, `!`) combine or negate TRUE/FALSE values that already exist.
- `&` means "and" — both must be TRUE
- `|` means "or" — at least one must be TRUE
- `!` means "not" — flips TRUE to FALSE and vice versa

A **logical mask** is NOT data. It has exactly two jobs:
- **Count the TRUEs:** `sum(mask)` tells you *how many* passed.
- **Extract the matching values:** `x[mask]` gives you the actual *values* that passed.

The trap is doing arithmetic directly on a mask — `mean(mask)`, `mask > 5`, `max(mask)`.
That almost always means you forgot to subset first.

The one exception: `mean(mask)` gives you the *proportion* of TRUEs (e.g. 0.6 = 60% passed).
This is useful when you actually want a percentage, but make sure that's what you intended.

The reflex to build:
> "Is this TRUE/FALSE or is it numbers? If TRUE/FALSE → either `sum()` it to count,
> or use `x[...]` to extract. Don't compute on it directly."

---

# Part A: Predict the output (no computer)

For each of these, **write down what you think R returns BEFORE running it.** Then run it and check. The point is to test your mental model, so don't skip the prediction step.

### Drill A.1

```{r}
x <- c(10, 20, 30, 40, 50)
mask <- x > 25

# Predict each of these before running:
# (a) What is `mask`? an vector with TRUEs and FALSEs
mask
# (b) What does sum(mask) give? tells you how many passed - 3
sum(mask)
# (c) What does x[mask] give? gives you the data that passed - 30, 40, and 50
x[mask]
# (d) What does mean(mask) give? (think carefully) the proportion that passed 0.6
mean(mask)
# (e) What does mean(x[mask]) give? the mean of the numbers that passed - 40
mean(x[mask])
```

> Write your five predictions as comments, THEN uncomment and run to check. Pay special attention to (d) vs (e) — they are very different, and the difference is the whole lesson.

### Drill A.2

```{r}
ages <- c(15, 22, 8, 67, 45, 30, 71)
adult <- ages >= 18

# Predict before running:
# (a) sum(adult)         -- ? 5
# (b) ages[adult]        -- ? 22, 67, 45, 30, 71
# (c) ages[!adult]       -- ? 15, 8 
# (d) sum(adult) / length(ages)   -- ? (what does this represent?) the proportion that are adults
sum(adult) / length(ages)
```

### Drill A.3

```{r}
v <- c(4, NA, 9, NA, 2, 7)
missing <- is.na(v)

# Predict before running:
# (a) missing            -- ? FALSE, TRUE, FALSE, TRUE, FALSE, FALSE
# (b) sum(missing)       -- ? 2
# (c) v[missing]         -- ? (what's the value at a missing position?) NA, NA
# (d) v[!missing]        -- ? 4, 9, 2, 7
# (e) sum(!missing)      -- ? (what does this count?) the number of non NA numbers - 4

```

---

# Part B: Fix the bug

Each of these snippets has a mask-vs-values bug (the same kind from your challenges). The intended result is described in the comment. Find the bug and fix it.

### Drill B.1

```{r}
scores <- c(82, 91, 67, 45, 78, 95, 58)

# GOAL: the average of all passing scores (>= 60)
# BUGGY VERSION:
passing <- scores >= 60
# avg_passing <- mean(passing)   # <-- this is wrong. why?
# it gives you the proportion that passed
# Write the corrected line:
avg_passing <- mean(scores[passing])
avg_passing
```

### Drill B.2

```{r}
temps <- c(68, 72, NA, 81, 90, NA, 75)

# GOAL: how many non-missing temperature readings are there?
# BUGGY VERSION:
# count <- sum(temps)            # <-- wrong (two problems here, actually)
# tries to add up the actual temperatures, but there is NA values within
# Write the corrected line:
count <- sum(!is.na(temps))
```

### Drill B.3

```{r}
prices <- c(12.50, 8.00, 25.00, 5.50, 18.00)

# GOAL: the highest price that is under $20
# BUGGY VERSION:
under_20 <- prices < 20
# highest <- max(under_20)       # <-- wrong. why?
# under_20 is a logical vector so there is no max
# Write the corrected line:
max(prices[under_20])
```

### Drill B.4

```{r}
mpg <- mtcars$mpg

# GOAL: the proportion of cars with mpg above 20 (a number between 0 and 1)
# BUGGY VERSION:
# prop <- mean(mpg > 20)         # <-- actually, is this wrong? think about it.
# gives the proportion of T/F under the logic of mpg > 20
# Decide whether this is correct. If it is, explain why in a comment.
# If not, fix it.
# i think that this is correct
```

> Drill B.4 is a trick question. `mean()` of a logical mask gives the *proportion of TRUEs*, because TRUE counts as 1 and FALSE as 0. So `mean(mpg > 20)` is actually a correct and idiomatic way to get a proportion. The lesson: computing on a mask isn't *always* wrong — but you have to know exactly what it means. `mean(mask)` = proportion; `sum(mask)` = count; `x[mask]` = values. Make sure that whenever you compute on a mask, you did it on purpose.

---

# Part C: Store-as-you-go

These drills target the habit of capturing intermediate values. Each function below is missing the storage step or computes the wrong thing because something wasn't saved.

### Drill C.1

```{r}
# GOAL: a function that returns the number of non-missing values in a column.
# BUGGY VERSION:
# count_present <- function(df, col) {
#   df[[col]]                       # this line does nothing
#   sum(!is.na(df[[col]]))
# }

# Rewrite it cleanly: store the column once as `x`, then work with `x`.
count_present <- function(df, col) {
  x <- df[[col]]
  sum(!is.na(x)) 
}

# Test:
count_present(airquality, "Ozone")   # should be 116

```

### Drill C.2

```{r}
# GOAL: a function that returns the mean of the values that are above the
# column's own median. (e.g., "average of the above-median cars' mpg")
# Build it step by step, storing each piece:

above_median_mean <- function(df, col) {
  x <- df[[col]] # Step 1: store the column as x
  median <- median(x, na.rm = T)          # Step 2: compute and store the median (NAs removed)
  above <- !is.na(x) & x > median # Step 3: build a mask: which values are above the median AND not NA?
  mean(x[above]) # Step 4: use the mask to extract values, then take their mean
}

# Test:
# above_median_mean(mtcars, "mpg")
above_median_mean(mtcars, "mpg")
```

### Drill C.3

```{r}
# GOAL: a function that takes a numeric vector and returns a named list with:
#   n_total    -- total number of elements
#   n_missing  -- number of NAs
#   n_present  -- number of non-NAs
#   mean_present -- mean of the non-NA values
# Store each intermediate result in its own variable before building the list.

describe <- function(x) {
  list(
    n_total = length(x),
    n_missing = sum(is.na(x)),
    n_present = sum(!is.na(x)),
    mean_present = mean(x, na.rm = T)
  )

}

# Test:
describe(c(5, NA, 8, 12, NA, 3))
# Expected: n_total = 6, n_missing = 2, n_present = 4, mean_present = 7

```

---

# Part D: Putting it together

One slightly bigger function that exercises both habits at once. Take your time and store every intermediate step.

### Drill D.1

```{r}
# GOAL: clean_above(df, col, cutoff)
# Returns a named list describing values in `col` that are above `cutoff`,
# after removing NAs:
#   n_dropped_na    -- how many NAs were removed
#   n_kept          -- how many non-NA values remain
#   n_above_cutoff  -- how many of the kept values are above cutoff
#   values_above    -- the actual vector of values above cutoff
#   mean_above      -- the mean of those above-cutoff values
#
# Build it with explicit, well-named intermediate variables.
# Remember: is.na() check first; mask to extract values, not to compute on.

clean_above <- function(df, col, cutoff) {
  x <- df[[col]]
  keep <- !is.na(x)
  clean <- x[keep]
  above <- clean > cutoff
  list(
    n_dropped_na = length(x) - length(x[keep]),
    n_kept = length(clean),
    n_above_cutoff = sum(above),
    values_above = clean[above], 
    mean_above = mean(clean[above])
  )
}

# Test:
result <- clean_above(airquality, "Ozone", 50)
result$n_above_cutoff   # how many ozone readings above 50
result$mean_above       # their average

```

> **Self-check before you call it done:** Walk through your function and for every variable ask: "Is this a mask (TRUE/FALSE) or values (numbers)?" Then confirm you only ever `sum()` masks or use them to subset — never `mean()` or `>` them by accident.
