---
title: "R Programming Drill: The Current Piece & Passing Functions"
author: "Tim Irwin"
date: "2026-05-20"
output:
  pdf_document: default
  html_document: default
---

# R Programming Drill: The Current Piece & Passing Functions

This is a focused drill, not a broad practice set. It targets the two specific habits that have shown up in your scratch work — not as separate weaknesses, but as one family of "what am I holding right now?" confusions.

**Pattern 1 — The current piece, not the whole thing.** Inside a loop or an apply function, the iterator variable (`x`, `sp`, `col`) is your handle on the *current* element. Referencing the full original object inside the loop body is almost always the bug.

**Pattern 2 — Pass a function, don't call one.** `lapply`, `tapply`, `mapply` want a *function* (a recipe they'll run later). Writing `sum(x)` hands them the result instead of the recipe. They want `function(x) sum(x)` (or just the name `sum`).

The exercises are short and repetitive on purpose. You're building a reflex.

```{r}
rm(list = ls())
```

> **The one question to ask, every time:** "At this exact spot in my code, am I holding the WHOLE object or the CURRENT PIECE? And am I handing over a FUNCTION or a RESULT?" Say it out loud if you have to. That sentence catches every bug in this drill.

## Datasets used

- `mtcars`, `iris`, `airquality`, `ChickWeight` — all built in.

---

# Part A: Spot the bug (no computer)

Each snippet below has exactly one bug, and it's always one of our two patterns. For each, **write a comment saying which pattern it is and what the fix is** — before running anything.

### Drill A.1

```{r}
# GOAL: mean Sepal.Width for each species
# results <- list()
# for (sp in unique(iris$Species)) {
#   results[[sp]] <- mean(iris$Sepal.Width)   # <-- bug?
# }

# Which pattern is broken? Write the fix as a comment, then write the
# corrected loop below and run it.

```

### Drill A.2

```{r}
# GOAL: the maximum of each numeric column of mtcars, as a list
# lapply(mtcars, max(x))   # <-- bug?

# Which pattern? Fix it below and run.

```

### Drill A.3

```{r}
# GOAL: for each Diet group (a data frame), the number of rows
# diet_list <- split(ChickWeight, ChickWeight$Diet)
# lapply(diet_list, function(d) nrow(ChickWeight))   # <-- bug?

# Which pattern? Fix it below and run.

```

### Drill A.4

```{r}
# GOAL: count, for each column, how many values exceed 100
# cols <- c("disp", "hp")
# mapply(function(col) sum(mtcars[[col]] > 100), cols)   # looks ok?

# This one actually runs. Is it correct? If yes, say why. If it's fragile,
# explain what would break it.

```

---

# Part B: Fill in the iterator

These all have the right structure. Your only job is to replace the WHOLE-object reference with the CURRENT-PIECE reference. The skeleton is done — just fix the body.

### Drill B.1

```{r}
# GOAL: mean Petal.Length per species, via a for loop
results <- list()
for (sp in unique(iris$Species)) {
  # FIX THIS LINE so it uses the current species `sp`, not the whole column:
  # results[[sp]] <- mean(iris$Petal.Length)
  
}
results
```

### Drill B.2

```{r}
# GOAL: for each diet's data frame, the mean weight
diet_list <- split(ChickWeight, ChickWeight$Diet)
# FIX the anonymous function so it uses the current data frame `d`:
# lapply(diet_list, function(d) mean(ChickWeight$weight))

```

### Drill B.3

```{r}
# GOAL: for each month's wind data, the range (max - min)
# tapply gives you the current group as `x` inside the anonymous function.
# FIX so it operates on `x`, not the whole column:
# tapply(airquality$Wind, airquality$Month, function(x) max(airquality$Wind) - min(airquality$Wind))

```

---

# Part C: Recipe, not result

These target Pattern 2. Each one calls a function where it should *pass* one. Rewrite each so it hands over a function (either an anonymous `function(x) ...` or just the bare function name).

### Drill C.1

```{r}
# GOAL: median of each numeric column of mtcars
# BROKEN: median() is being called, not passed
# lapply(mtcars, median(mtcars))

# Two valid fixes exist — write BOTH:
# (a) pass the bare function name
# (b) pass an anonymous function

```

### Drill C.2

```{r}
# GOAL: mean Wind per month
# BROKEN: mean() is called immediately
# tapply(airquality$Wind, airquality$Month, mean(airquality$Wind))

# Write the corrected version (passing the function, not calling it):

```

### Drill C.3

```{r}
# GOAL: for paired vectors, raise each base to its matching power
bases  <- c(2, 3, 4, 5)
powers <- c(5, 4, 3, 2)
# BROKEN: the operation is computed instead of wrapped in a function
# mapply(bases ^ powers, bases, powers)

# Write the corrected version using function(b, p) ... :

```

---

# Part D: Build it from a blank line

No skeleton this time. Write each from scratch, narrating the "whole vs piece / function vs result" question in comments as you go.

### Drill D.1

```{r}
# Using a for loop, build a named list giving the mean Sepal.Length
# for each iris species. Before each line, jot a comment: am I about to
# reference the whole column or the current species?

```

### Drill D.2

```{r}
# Using tapply with an anonymous function, return for each iris species a
# named vector c(mean = ..., sd = ...) of Sepal.Length.
# (This is the "anonymous function returning a named vector" pattern you
#  already nailed in the loop functions practice — prove to yourself it
#  wasn't a fluke.)

```

### Drill D.3

```{r}
# You have two parallel vectors:
species_list <- c("setosa", "versicolor", "virginica")
cutoffs      <- c(5.0, 6.0, 6.5)
# For each (species, cutoff) pair, count how many irises of that species
# have Sepal.Length above that cutoff. Use mapply with an anonymous
# function of two arguments. Result: a named vector of three counts.

```

---

# Part E: One real, messy task

Pull it together on something slightly bigger. No hints about which pattern is where — just stay disciplined with the question.

### Drill E.1

```{r}
# For each cylinder group in mtcars (4, 6, 8), produce a named list where
# each element is itself a named vector containing:
#   n        -- number of cars in that group
#   avg_mpg  -- their average mpg
#   avg_hp   -- their average hp
#
# Use split() to break mtcars into groups, then lapply() with an anonymous
# function that receives each group's DATA FRAME and pulls what it needs
# from THAT data frame (not from mtcars).
#
# Self-check before running: inside your anonymous function, every reference
# should be to the parameter (e.g. `g`), never to `mtcars`.

```

> **After you finish E.1**, reread your anonymous function and circle every object name. If any of them says `mtcars` instead of your parameter, that's Pattern 1 trying to sneak back in. That circling habit — done for ten seconds before you run — is what turns this from a thing you catch *after* an error into a thing you catch *before*.
