---
title: "R Programming Practice (Part 7): Debugging"
author: "Tim Irwin"
date: "2026-05-20"
output:
  pdf_document: default
  html_document: default
---

# R Programming Practice (Part 7): Debugging

This practice set covers debugging from *R Programming for Data Science* by Roger D. Peng (pages 94–101):

1. The three signals: `message`, `warning`, `error`
2. Reading and interpreting error messages
3. A diagnostic process for figuring out what went wrong
4. R's debugging tools: `traceback()`, `debug()`, `browser()`, `recover()`
5. Defensive coding: anticipating problems before they happen

Each section starts with a **lesson** that teaches the concept with a worked example, followed by **exercises**. Debugging is a slightly different kind of skill from the earlier practice sets — instead of writing new code, a lot of these exercises hand you *broken* code and ask you to diagnose and fix it. That's deliberate: reading and fixing broken code is most of what real programming actually is.

You may want to start with:

```{r}
rm(list = ls())
```

> **Tip:** When something breaks, resist the urge to randomly change code until it works. Debugging is detective work: figure out what you expected, what actually happened, and where they diverged — *then* fix. Peng's book ends the chapter with a line worth remembering: "Debugging tools are not a substitute for thinking!"

## Datasets used in this practice

- `airquality` — Daily air quality measurements in New York, May–September 1973.
- `mtcars` — Fuel consumption and 10 aspects of design/performance for 32 cars.
- `iris` — Measurements of 150 iris flowers across 3 species.

---

# 1. The Three Signals: message, warning, error

## Lesson 1.1: R has three levels of "something happened"

When R wants to tell you something, it uses one of three levels, in increasing severity:

- **message** — a gentle notification. Execution continues normally. Produced by `message()`.
- **warning** — "something looks off, but I'll keep going." Execution continues, but you should pay attention. Produced by `warning()`.
- **error** — "I can't continue." Execution stops immediately. Produced by `stop()`.

Only an **error** is fatal. Messages and warnings let the code keep running.

Here's each one in action:

```{r}
# A message — informational, execution continues
demo_message <- function() {
  message("Just so you know: starting the calculation")
  42
}
demo_message()
```

```{r}
# A warning — something's off, but R returns a value anyway
sqrt(-4)   # can't take sqrt of a negative, so R warns and returns NaN
```

```{r error=TRUE}
# An error — execution halts (error=TRUE lets this doc keep knitting)
demo_error <- function() {
  stop("I cannot proceed")
  42   # this line never runs
}
demo_error()
```

> **Tip:** In R Markdown, a code chunk that throws an error normally stops the whole document from knitting. Adding `error=TRUE` to the chunk header (like `{r error=TRUE}`) tells R to show the error and keep going. That's how this document can demonstrate errors without breaking.

### Exercise 1.1

Write a function `check_value(x)` that:

- Uses `message()` to print `"Checking the value..."` (informational)
- If `x` is negative, uses `warning()` to warn `"Value is negative"` but still returns `x`
- If `x` is exactly 0, uses `stop()` to halt with `"Value cannot be zero"`
- Otherwise just returns `x`

Test it with `check_value(5)`, `check_value(-3)`, and `check_value(0)`. (Put the zero test in a chunk with `error=TRUE` so the doc still knits.)

```{r}
# Your code here

```

```{r error=TRUE}
# Test the error case here

```

---

## Lesson 1.2: Why warnings matter — they're easy to miss

The danger of warnings is that your code *keeps running*, so you might not notice something went wrong until much later. Watch this:

```{r}
ages <- c("34", "29", "forty", "51")
numeric_ages <- as.numeric(ages)   # "forty" can't convert
numeric_ages
mean(numeric_ages)   # the NA silently poisons the result
```

The `as.numeric()` call warned that NAs were introduced (because `"forty"` isn't a number), but it didn't stop. Then `mean()` returned `NA` because of that one bad value. If you weren't watching for the warning, you might just see `NA` and be confused.

### Exercise 1.2

The vector below is supposed to contain temperatures, but someone typed one of them as a word. Convert it to numeric, observe the warning, then compute the mean correctly (ignoring the value that couldn't convert).

```{r}
temps <- c("68", "72", "seventy", "81", "75")
# Your code here: convert to numeric, then compute the mean ignoring NAs

```

---

# 2. Reading Error Messages

## Lesson 2.1: Error messages tell you where and what

A good chunk of debugging is just *reading the error carefully*. R error messages usually tell you two things: where the problem happened and what kind of problem it was. Consider:

```{r error=TRUE}
total <- sum(values)   # `values` was never created
```

The message — `object 'values' not found` — is precise. It's not a mysterious failure; R is telling you exactly what's wrong: you referenced a name that doesn't exist. Maybe a typo, maybe you forgot to create it, maybe you're running lines out of order.

Here's another extremely common one:

```{r error=TRUE}
classify_temp <- function(temp) {
  if (temp > 80) {
    "hot"
  } else {
    "not hot"
  }
}

# Now apply it to a whole column — what happens?
classify_temp(airquality$Temp)
```

The error is `the condition has length > 1`. This is one of the most common R errors, and you've actually run into its root cause before: `if` expects a **single** TRUE/FALSE, but `airquality$Temp > 80` is a whole *vector* of TRUE/FALSEs. The function was written for one value at a time, but you fed it a vector.

> **Tip:** "the condition has length > 1" almost always means you put a vector where `if` expected a single value. The fix is usually to either vectorize the function (loop over elements, or use `ifelse()`) or check `length()` and handle vectors explicitly.

### Exercise 2.1

The error below is one of the most common in R. Read the error message, explain in a comment what's actually wrong, and then fix it so the function correctly classifies the *whole* `mtcars$mpg` vector. Hint: `ifelse()` is a vectorized version of if/else that works on whole vectors.

```{r error=TRUE}
rate_mpg <- function(x) {
  if (x > 25) {
    "good"
  } else {
    "poor"
  }
}
rate_mpg(mtcars$mpg)   # this errors
```

```{r}
# Your fixed version here — should return a vector of "good"/"poor"

```

---

## Lesson 2.2: The NA-in-a-condition error

Here's a function that looks fine but breaks on real data. This is Peng's classic example, recreated with different code:

```{r error=TRUE}
describe_reading <- function(x) {
  if (x > 100) {
    "high"
  } else {
    "normal"
  }
}

# Works on a clean value:
describe_reading(150)
```

```{r error=TRUE}
# But airquality$Ozone has NAs. Grab one and try:
describe_reading(airquality$Ozone[5])   # row 5 is NA
```

The error — `missing value where TRUE/FALSE needed` — happens because `NA > 100` evaluates to `NA`, and `if` can't decide what to do with `NA`. It needs a definite TRUE or FALSE.

The fix is to check for NA *first*, before any other comparison (you've done this before in your functions practice):

```{r}
describe_reading2 <- function(x) {
  if (is.na(x)) {
    "missing"
  } else if (x > 100) {
    "high"
  } else {
    "normal"
  }
}

describe_reading2(airquality$Ozone[5])   # now returns "missing"
describe_reading2(150)
```

### Exercise 2.2

The function below crashes when it hits an NA. Reproduce the error first (in the `error=TRUE` chunk), then write a fixed version that returns `"unknown"` for missing values. Test the fix on `airquality$Solar.R[5]` (which is NA) and on a normal value.

```{r error=TRUE}
solar_level <- function(x) {
  if (x > 200) {
    "strong"
  } else {
    "weak"
  }
}
solar_level(airquality$Solar.R[5])   # reproduce the error
```

```{r}
# Your fixed version here

```

---

# 3. A Diagnostic Process

## Lesson 3.1: The six questions

Peng lays out a set of questions to ask whenever something breaks. They sound obvious, but actually asking them — out loud or in writing — is what separates efficient debugging from flailing:

1. What was your input? How did you call the function?
2. What were you expecting? (Output, messages, results?)
3. What did you actually get?
4. How does what you got differ from what you expected?
5. Were your expectations correct in the first place?
6. Can you reproduce the problem exactly?

Question 5 is the sneaky one. Sometimes the code is fine and your *expectation* was wrong. And question 6 matters because a bug you can't reproduce is a bug you can't fix.

Here's the process in action. Suppose you write this and get a surprising result:

```{r}
# I expected the mean mpg of 6-cylinder cars
six_cyl_mean <- mean(mtcars$mpg[mtcars$cyl = 6])
```

Wait — that should have errored. Let me show the real version:

```{r error=TRUE}
six_cyl_mean <- mean(mtcars$mpg[mtcars$cyl = 6])
```

The error points at the `=` sign. Working through the questions: my *input* was a subsetting expression; I *expected* the mean of 6-cylinder cars; I *got* a syntax error about an unexpected `=`. The divergence: I used `=` (assignment) where I meant `==` (comparison). Classic typo.

```{r}
# Corrected:
six_cyl_mean <- mean(mtcars$mpg[mtcars$cyl == 6])
six_cyl_mean
```

### Exercise 3.1

The code below is *supposed* to compute the average `Sepal.Length` of setosa irises, but it returns something wrong (not an error — a wrong answer, which is more dangerous). Work through Peng's questions: what did you expect, what did you get, and where's the divergence? Write your diagnosis as comments, then fix it.

```{r}
# This runs without error but gives the wrong answer:
wrong_answer <- mean(iris$Sepal.Length[iris$Species == "Setosa"])
wrong_answer
# Hint: the result is NaN. Why? What's different about "Setosa" vs the actual data?

```

```{r}
# Your corrected version here

```

> **Tip:** R is case-sensitive. The species in `iris` is `"setosa"` (lowercase), not `"Setosa"`. When a comparison matches *nothing*, you get an empty vector, and `mean()` of an empty vector is `NaN`. This is exactly why question 5 — "were my expectations correct?" — matters. The code did exactly what you told it; you just told it the wrong string.

---

# 4. Debugging Tools

## Lesson 4.1: The humble print/cat statement

Before the fancy tools, know this: **inserting `print()` or `cat()` statements is the most common debugging technique in the world**, and there's no shame in it. When you're not sure what's happening inside a function, print things out to see.

```{r}
# A function that's misbehaving — let's add print statements to watch it work
buggy_sum <- function(x) {
  total <- 0
  for (i in seq_along(x)) {
    cat("i =", i, " x[i] =", x[i], " total so far =", total, "\n")  # debugging line
    total <- total + x[i]
  }
  total
}

buggy_sum(c(5, 10, NA, 20))
```

The printed trace shows you exactly where things go wrong: as soon as it hits the NA, `total` becomes NA and stays that way. The print statements made the invisible visible.

### Exercise 4.1

The function below is supposed to count how many values in a vector are above a threshold, but it always returns 0. Add `cat()` or `print()` statements inside the loop to watch what's happening, figure out the bug, and fix it.

```{r}
count_above <- function(x, threshold) {
  count <- 0
  for (i in seq_along(x)) {
    if (x[i] > threshold) {
      count + 1   # <-- something's wrong here
    }
  }
  count
}

count_above(c(10, 20, 30, 40), 15)   # should return 3, but returns 0
```

```{r}
# Your debugged version here (add print statements first to find it, then fix)

```

> **Tip:** The bug in 4.1 is the same *category* as one you hit before — a result that's computed but never stored. `count + 1` calculates a value and throws it away. It needs to be `count <- count + 1`. Watching `count` with a print statement makes this obvious: it never changes.

---

## Lesson 4.2: traceback()

When an error happens deep inside nested function calls, it's not always clear *which* function actually failed. `traceback()` shows you the chain of calls that led to the error — the "call stack."

You call `traceback()` immediately after an error occurs (it must be the very next thing you do). Imagine three functions that call each other:

```{r error=TRUE}
inner <- function(x) {
  x + missing_object   # this will error
}
middle <- function(x) {
  inner(x)
}
outer <- function(x) {
  middle(x)
}

outer(5)   # errors deep inside `inner`
```

After running this in your console, you would type `traceback()` and see something like:

```
3: inner(x)
2: middle(x)
1: outer(5)
```

Read it bottom to top: you called `outer`, which called `middle`, which called `inner`, and the error happened in `inner`. Now you know exactly where to look.

> **Note:** `traceback()` is interactive — it works in the RStudio console, not inside a knitted document. So for this exercise, you'll run the code in your console and call `traceback()` there.

### Exercise 4.2

In your console (not the knitted doc), define these three functions and call the outer one, then run `traceback()` to confirm which function the error originates in. Write a comment describing what the traceback showed you.

```{r eval=FALSE}
step_a <- function(v) {
  mean(v) + undefined_thing   # the bug lives here
}
step_b <- function(v) {
  step_a(v)
}
step_c <- function(v) {
  step_b(v)
}

step_c(mtcars$mpg)
# Then run traceback() in the console and note what it showed
```

```{r}
# Write your observation as a comment here:

```

---

## Lesson 4.3: browser()

`browser()` is the most powerful tool here. Inserting `browser()` into a function pauses execution at that exact spot and drops you into an interactive session where you can inspect every variable, run code, and step forward line by line.

Inside the browser, a few key commands:

- `n` — execute the current line and move to the next
- `c` — continue running until the function finishes (or errors)
- `Q` — quit the browser
- typing any variable name — prints its current value

```{r eval=FALSE}
# Example (run this in your console, not the knitted doc):
inspect_me <- function(x) {
  doubled <- x * 2
  browser()              # execution pauses here
  result <- doubled + 10
  result
}

inspect_me(5)
# At the Browse> prompt, type `doubled` to see its value (10),
# then `n` to step to the next line, then `c` to finish.
```

> **Note:** Like `traceback()`, `browser()` is interactive and only works in a live R session, not in a knitted document. Try it in your console to get a feel for stepping through code.

### Exercise 4.3

In your console, take the `count_above` function you fixed in Exercise 4.1, insert a `browser()` just inside the loop, and step through a couple of iterations using `n`. At each pause, inspect the value of `count` and `x[i]`. Write a comment describing what you observed about how `count` changes (or doesn't) across iterations.

```{r}
# Write your observation as a comment here:

```

---

# 5. Defensive Coding

## Lesson 5.1: Catching problems before they happen

The best debugging is the kind you never have to do, because you wrote the function to fail loudly and early when given bad input. You've already practiced this with `stop()` in the validation challenge. Here it's framed as a debugging strategy.

Consider a function that assumes its input is a single positive number. Without checks, bad input produces confusing downstream errors. With checks, it fails immediately with a clear message:

```{r error=TRUE}
safe_sqrt <- function(x) {
  if (length(x) != 1) {
    stop("x must be a single number, but it has length ", length(x))
  }
  if (!is.numeric(x)) {
    stop("x must be numeric, but it is ", class(x))
  }
  if (x < 0) {
    warning("x is negative; result will be NaN")
  }
  sqrt(x)
}

safe_sqrt(16)        # fine
```

```{r error=TRUE}
safe_sqrt(c(1, 2, 3))   # fails early with a clear message
```

```{r}
safe_sqrt(-4)        # warns but proceeds
```

Notice the error message uses the actual length and class, so the caller knows exactly what they did wrong. That's the difference between a helpful error and a cryptic one.

### Exercise 5.1

Write a defensive function `safe_divide(a, b)` that divides `a` by `b`, but:

- Stops with a clear message if either `a` or `b` is not numeric (mention which one and its actual class)
- Stops with a clear message if `b` is 0 (dividing by zero)
- Otherwise returns `a / b`

Test it with a valid call, a divide-by-zero call, and a non-numeric call (each problematic one in an `error=TRUE` chunk).

```{r}
# Your code here

```

```{r error=TRUE}
# Test the divide-by-zero case

```

```{r error=TRUE}
# Test the non-numeric case

```

---

## Lesson 5.2: The vectorization bug, revisited

Recall the "condition has length > 1" error from Lesson 2.1. A robust function should either handle vectors properly or refuse them with a clear message. Here's a function that detects the problem and tells the caller exactly what to do:

```{r error=TRUE}
grade_one <- function(score) {
  if (length(score) > 1) {
    stop("grade_one() takes a single score. For a vector, use sapply(scores, grade_one).")
  }
  if (score >= 60) "pass" else "fail"
}

grade_one(75)            # works
```

```{r error=TRUE}
grade_one(c(75, 50, 90)) # fails with a helpful instruction
```

```{r}
# And here's how the caller would apply it to a vector, as the message suggests:
sapply(c(75, 50, 90), grade_one)
```

### Exercise 5.2

Take the `solar_level` function you fixed in Exercise 2.2 (the one that handles NA). Add a length check at the very top that stops with a helpful message if given a vector longer than 1. Then demonstrate both: the error when you pass the whole `airquality$Solar.R` column, and the correct result when you use `sapply()` to apply it across the column. (Use `head()` to keep the `sapply` output short.)

```{r}
# Your code here — the length-checked version of solar_level

```

```{r error=TRUE}
# Demonstrate the error by passing the whole column

```

```{r}
# Demonstrate the correct vectorized usage with sapply (use head() to keep it short)

```

---

# Mini Challenges

These combine diagnosis and fixing. Each one hands you broken code — your job is to find and fix the bug, ideally using the tools and process from the lessons.

## Challenge 1: Three bugs in one function

The function below is meant to take a numeric vector and return the mean of all values strictly above the average. It has **three** separate bugs. Find and fix all of them. (Hint: use print statements, and watch for the mask-vs-values trap.)

```{r error=TRUE}
above_avg_mean <- function(x) {
  avg <- mean(x)                    # bug-prone if there are NAs
  mask <- x > avg
  result <- mean(mask)              # is this the mean we want?
  reslut                            # returned value
}

above_avg_mean(c(4, 8, 15, 16, 23, 42))
```

```{r}
# Your fully corrected version here. It should return the mean of the
# values above the average. Test it on c(4, 8, 15, 16, 23, 42).

```

---

## Challenge 2: The silent wrong answer

This function runs without any error or warning, but gives a wrong result. It's supposed to return the number of `airquality` rows where `Temp` is above 85. Diagnose why it's wrong using Peng's questions, then fix it.

```{r}
hot_days <- function(df) {
  mask <- df$Temp > 85
  length(mask)        # this returns something, but is it the count of hot days?
}

hot_days(airquality)
# Hint: length(mask) vs sum(mask) — what's the difference?

```

```{r}
# Your corrected version here

```

---

## Challenge 3: Reproduce, then fix

The function below works sometimes and fails other times. Your job: first figure out *what inputs make it fail* (reproduce the problem reliably), then fix it so it works on all of them.

```{r error=TRUE}
risk_category <- function(ozone) {
  if (ozone > 100) {
    "hazardous"
  } else if (ozone > 50) {
    "elevated"
  } else {
    "acceptable"
  }
}

# These work:
risk_category(120)
risk_category(30)
```

```{r error=TRUE}
# Find an input from airquality$Ozone that breaks it, and reproduce the error:
# (Hint: what kind of values does airquality$Ozone contain that 120 and 30 don't?)

```

```{r}
# Your fixed version here that handles ALL values in airquality$Ozone,
# then apply it across the whole column with sapply() and table() the results.

```

---

## Challenge 4: Debugging a pipeline

This multi-step function chains several operations. It errors somewhere in the middle. Use print statements (or `browser()` in your console) to find which step fails and why, then fix it.

```{r error=TRUE}
summarize_group <- function(df, group_col, value_col) {
  groups <- unique(df[[group_col]])
  results <- numeric(length(groups))
  for (i in seq_along(groups)) {
    g <- groups[i]
    subset_rows <- df[df[[group_col]] == g, ]
    results[i] <- mean(subset_rows[value_col], na.rm = TRUE)   # bug is around here
  }
  names(results) <- groups
  results
}

# Should give average mpg for each cylinder count:
summarize_group(mtcars, "cyl", "mpg")
```

```{r}
# Your corrected version here.
# Hint: subset_rows[value_col] vs subset_rows[[value_col]] — one gives a
# data frame, the other gives the vector you actually want to average.

```

---

## Challenge 5: Write it defensively from scratch

Now flip the script: instead of fixing broken code, write robust code. Write `column_zscore(df, col)` that returns the z-scores of a column (each value minus the column mean, divided by the standard deviation). But make it defensive:

- Stop if `df` isn't a data frame (clear message)
- Stop if `col` isn't a single character string
- Stop if `col` isn't a column of `df` (name the bad column)
- Stop if the column isn't numeric (name the column and its class)
- Handle NAs correctly in the mean and sd calculations
- Return the vector of z-scores

Test it on `mtcars` with `"mpg"` (valid), and trigger at least two of the error conditions in `error=TRUE` chunks.

> **Tip:** A z-score is `(x - mean) / sd`. Because R is vectorized, once you have the cleaned column `x`, the whole calculation is one line: `(x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)`. The defensive checks are the substance of this challenge.

```{r}
# Your code here

```

```{r error=TRUE}
# Trigger an error condition

```

```{r error=TRUE}
# Trigger a different error condition

```
