---
title: "R Programming Practice (Part 5): Functions — Additional Challenges"
author: "Tim Irwin"
date: "2026-04-27"
output:
  pdf_document: default
  html_document: default
---

# R Programming Practice (Part 5): Functions — Additional Challenges

These are bonus challenges that build on the Functions practice (Part 5). They reuse some of the functions you already wrote — `vec_summary`, `classify_all`, `analyze_column` — and extend them in new directions. Each challenge introduces a new wrinkle: input validation, multiple NULL defaults, building larger pipelines, and more.

The challenges are roughly ordered from easier to harder. There are usually multiple right approaches — focus on writing clear, working code.

```{r}
rm(list = ls())
```

## Datasets used

- `airquality` — Daily air quality measurements in New York, May–September 1973.
- `mtcars` — Fuel consumption and 10 aspects of design/performance for 32 cars.
- `iris` — Measurements of 150 iris flowers across 3 species.

---

## Challenge 1: Letter grade converter

Write `letter_grade(score)` that takes a single numeric score (0–100) and returns the corresponding letter:

- 90 or above → `"A"`
- 80–89 → `"B"`
- 70–79 → `"C"`
- 60–69 → `"D"`
- Below 60 → `"F"`

Then write `letter_grade_all(scores)` that vectorizes it — takes a vector of scores and returns a vector of letters. Apply it to `c(95, 73, 88, 51, 67, 100, 79)`.

```{r}
### got it right
letter_grade <- function(score) {
  if (score >= 90) {
    "A"
  } else if (score >= 80 & score <= 89) {
    "B"
  } else if (score >= 70 & score <= 79) {
    "C"
  } else if (score >= 60 & score <=69) {
    "D"
  } else {
    "F"
  }
}

letter_grades_all <- function(scores) {
  results <- character(length(scores)) #pre-allocate result vector, same length
  # as vector input
  for (i in seq_along(scores)) { # loops through index positions (1, 2, 3...) 
    # of the scores vector
    results[i] <- letter_grade(scores[i]) # grabs the ith score and passes it to
    # letter_grade function and stores the result in the ith position
  }
  results
}
letter_grades_all(c(95, 73, 88, 51, 67, 100, 79))
```

---

## Challenge 2: Tax bracket calculator

Write `calculate_tax(income)` that computes total tax owed using these (simplified) progressive brackets:

- The first \$10,000 of income is taxed at 10%
- Income from \$10,001 to \$50,000 is taxed at 20%
- Income above \$50,000 is taxed at 30%

So for an income of \$60,000, you'd owe: `10000 * 0.10 + 40000 * 0.20 + 10000 * 0.30 = 12000`.

The trick is that *each portion* of income is taxed at a different rate — not the whole amount at the highest rate. Use `if/else if` to handle the three cases.

Test with incomes of `5000`, `30000`, `60000`, and `100000`.

> **Tip:** Think about what tax is owed in each case and break the math into pieces. For income above $50k, you owe the full tax on the first $10k bracket, the full tax on the $10k–$50k bracket, and 30% on whatever is above $50k.

```{r}
#### got it right
calculate_tax <- function(income) {
  if (income <= 10000) {
    liability <- income * 0.1
  } else if (income > 10000 & income <= 50000) {
    remainder <- income - 10000
    liability <- (10000 * 0.1) + (remainder * 0.2)
  } else {
    surplus <- income - 50000
    liability <- (10000 * 0.1) + (40000 * 0.2) + (surplus * 0.3)
  }
  liability
}

calculate_tax(5000)
calculate_tax(30000)
calculate_tax(60000)
calculate_tax(100000)
```

---

## Challenge 3: Range filter with two bounded defaults

Write `filter_range(df, col, low = NULL, high = NULL)` that returns the subset of rows where the value in `col` is between `low` and `high` (inclusive).

The twist: **either or both** bounds can be `NULL`. Handle all four cases:

- Both `NULL` → return all rows (no filtering)
- Only `low = NULL` → return rows where `col <= high`
- Only `high = NULL` → return rows where `col >= low`
- Both supplied → return rows where `low <= col <= high`

Test on `mtcars` with `"mpg"`:

1. `filter_range(mtcars, "mpg", low = 20, high = 25)` — cars with mpg between 20 and 25
2. `filter_range(mtcars, "mpg", low = 25)` — cars with mpg 25 or higher (no upper bound)
3. `filter_range(mtcars, "mpg", high = 15)` — cars with mpg 15 or lower (no lower bound)

```{r original3}
##### ORIGINAL
# filter_range <- function(df, col, low = NULL, high = NULL) {
#   if (is.null(low) & is.null(high)) {
#     df[df[[col]]]
#   } else if (is.null(low) & !is.null(high)) {
#     df[df[[col]] <= high, ]
#   } else if (!is.null(low) & is.null(high)) {
#     df[df[[col]] >= low, ]
#   } else if (!is.null(low) & !is.null(high)) {
#     df[df[[col]] >= low, ] & df[df[[col]] <= high, ]
#   }
# }

```
```{r corrected3}
##### CORRECTED
# function takes a data frame, column name, and optional low/high bounds
# low and high default to NULL so the caller can omit either or both
filter_range <- function(df, col, low = NULL, high = NULL) {
  # is.null() checks if the argument was left empty
  # & means both conditions must be TRUE
  if (is.null(low) & is.null(high)) {
    df # no filtering needed, return the whole data frame as-is
  } else if (is.null(low) & !is.null(high)) {
    # !is.null(high) means high was provided
    # df[[col]] pulls the column as a vector
    # <= high creates a logical vector (TRUE/FALSE for each row)
    # df[logical_vector, ] keeps only rows where TRUE
    # the comma is needed to say "all columns"
    df[df[[col]] <= high, ]
  } else if (!is.null(low) & is.null(high)) {
    # same idea but filtering for values >= low
    df[df[[col]] >= low, ]
  } else if (!is.null(low) & !is.null(high)) {
    # & combines two conditions: must be >= low AND <= high
    # both conditions go inside the same brackets
    df[df[[col]] >= low & df[[col]] <= high, ]
  }
}

filter_range(mtcars, "mpg", low = 20, high = 25)
filter_range(mtcars, "mpg", low = 25)
filter_range(mtcars, "mpg", high = 15)
```

---

## Challenge 4: Input validation

Write `safe_column_mean(df, col)` that returns the mean of `col` (NAs removed), but **first validates the inputs**:

1. If `df` is not a data frame, stop with `"df must be a data frame"`.
2. If `col` is not a single character string, stop with `"col must be a single column name"`.
3. If `col` is not actually a column of `df`, stop with a message like `"Column 'banana' not found in df"` (showing the actual column name they passed).
4. If the column exists but isn't numeric, stop with `"Column 'X' must be numeric"`.

If all checks pass, return the mean.

Test it with valid inputs (`safe_column_mean(mtcars, "mpg")`) and at least two invalid inputs to confirm the right errors fire.

> **Tip:** `is.data.frame(df)` checks if something is a data frame. `is.character(col) && length(col) == 1` checks for a single string. `col %in% colnames(df)` checks if the column name exists. `paste0()` lets you build messages with the actual value the caller passed.

```{r original4, error=TRUE}
####### MY ORIGINAL
# safe_column_mean <- function(df, col) {
#   if (!is.data.frame(df)) {
#       stop("df must be a dataframe")
#   }
#   if (!is.character(col) && !length(col) == 1) {
#     stop("col must be a single column name")
#   }
#   if (!col %in% colnames(df)) {
#     stop("Column", paste0(col), "not found in the df")
#   }
#   if (!is.numeric(df)) {
#     stop("Column", paste0(col), "must be numeric")
#   }
#   mean(col, na.rm = TRUE)
# }
```

```{r corrected4, error=TRUE}
###### CORRECTED
safe_column_mean <- function(df, col) {
  # check 1: is.data.frame() returns TRUE/FALSE
  # ! negates it — so this reads "if df is NOT a data frame, stop"
  if (!is.data.frame(df)) {
      stop("df must be a dataframe")
  }
  # check 2: col should be a character string AND length 1 (a single name, not a vector of names)
  # && is used here instead of & — && is better for single TRUE/FALSE checks in if statements
  # wrap the whole condition in !() to negate it — NOT just the first part
  # WRONG: !is.character(col) && !length(col) == 1 — this means BOTH must fail
  # RIGHT: !(is.character(col) && length(col) == 1) — this means if EITHER fails, stop
  if (!(is.character(col) && length(col) == 1)) {
    stop("col must be a single column name")
  }
  # check 3: %in% checks if col is one of the column names in df
  # need parentheses around the whole thing: !(col %in% colnames(df))
  # WRONG: !col %in% colnames(df) — the ! only negates col, not the whole expression
  if (!(col %in% colnames(df))) {
    stop("Column", paste0(col), "not found in the df")
  }
  # check 4: check the actual column, not the whole data frame
  # df[[col]] pulls the column as a vector — then is.numeric() tests it
  # WRONG: !is.numeric(df) — this checks the whole data frame
  # RIGHT: !is.numeric(df[[col]]) — this checks just the column
  if (!is.numeric(df[[col]])) {
    stop("Column", paste0(col), "must be numeric")
  }
  # all checks passed — safe to compute the mean
  # must use df[[col]] to get the actual values, not just col (which is a string)
  # WRONG: mean(col, na.rm = TRUE) — col is just the name "mpg", not the data
  # RIGHT: mean(df[[col]], na.rm = TRUE) — this gets the actual numbers
  mean(df[[col]], na.rm = TRUE)
}


safe_column_mean(mtcars, "mpg")          # should return 20.09062
safe_column_mean("not a df", "mpg")      # should stop: df must be a dataframe
safe_column_mean(mtcars, 123)            # should stop: col must be a single column name
safe_column_mean(mtcars, "banana")       # should stop: Column banana not found
safe_column_mean(iris, "Species")        # should stop: Column Species must be numeric
```

---

## Challenge 5: Flexible conditional counter

Write `count_matching(x, threshold, comparison = "greater")` that counts how many values in `x` match a condition relative to `threshold`. The `comparison` argument controls what kind of comparison:

- `"greater"` → count values > threshold
- `"less"` → count values < threshold
- `"equal"` → count values == threshold
- `"at_least"` → count values >= threshold
- `"at_most"` → count values <= threshold

If `comparison` is anything else, stop with an informative error.

Handle NAs by always ignoring them (don't count NAs as matches, and don't error out on them).

Test all five comparisons on `mtcars$mpg` with `threshold = 20`. You should get five different counts.

```{r original 5}
##### ORIGINAL
# count_matching <- function(x, threshold, comparison = "greater") {
#   count <- 0                                            
#   if (comparison == "greater") {
#     count > threshold
#   } else if (comparison == "less") {
#     count < threshold
#   } else if (comparison == "equal") {
#     count == threshold
#   } else if (comaparison == "at_least") {
#     count >= threshold
#   } else if (comparison == "at_most") {
#     count <= threshold
#   } else {
#     stop("comparison error")
#   }
# 
# }

```
```{r corrected5}
##### CORRECTED
# takes a vector x, a threshold value, and a comparison type (defaults to "greater")
count_matching <- function(x, threshold, comparison = "greater") {
  # check what type of comparison the caller wants
  # each branch compares every element of x against threshold (vectorized)
  # x > threshold returns a logical vector (TRUE/FALSE for each element)
  # sum() counts the TRUEs — since TRUE = 1 and FALSE = 0
  # na.rm = TRUE tells sum() to skip any NAs instead of returning NA
  if (comparison == "greater") {
    sum(x > threshold, na.rm = TRUE)
  } else if (comparison == "less") {
    sum(x < threshold, na.rm = TRUE)
  } else if (comparison == "equal") {
    sum(x == threshold, na.rm = TRUE)
  } else if (comparison == "at_least") {
    sum(x >= threshold, na.rm = TRUE)
  } else if (comparison == "at_most") {
    sum(x <= threshold, na.rm = TRUE)
  } else {
    # if comparison doesn't match any valid option, stop with an error
    # this catches typos or invalid inputs like "bigger" or "more"
    stop("comparison error")
  }
}

count_matching(mtcars$mpg, 20, "greater")
count_matching(mtcars$mpg, 20, "less")
count_matching(mtcars$mpg, 20, "equal")
count_matching(mtcars$mpg, 20, "at_least")
count_matching(mtcars$mpg, 20, "at_most")
```

---

## Challenge 6: Two-column relationship

Write `column_relationship(df, col_a, col_b)` that takes a data frame and two column names, and returns a named list summarizing the relationship between them:

- `correlation` — the correlation between the two columns (use `cor()` with `use = "complete.obs"` to handle NAs)
- `mean_a` — the mean of `col_a` (NAs removed)
- `mean_b` — the mean of `col_b` (NAs removed)
- `mean_difference` — `mean_a - mean_b`
- `n_complete` — the number of rows where both columns are non-missing

Test it on `mtcars` with `"mpg"` and `"hp"`. You should see a strong negative correlation (heavier-engined cars get worse mpg).

```{r original6}
##### Original
# column_relationship <- function(df, col_a, col_b) {
#   list(
#     correlation = cor(df[[col_a]], df[[col_b]], use = "complete.obs"),
#     mean_a = mean(df[[col_a]], na.rm = T),
#     mean_b = mean(df[[col_b]], na.rm = T),
#     mean_difference = mean_a - mean_b,
#     n_complete = sum(!(is.na(df[[col_a]] & df[[col_b]])))
#   )
# }
```

```{r corrected6}
##### CORRECTED
# takes a data frame and two column names, returns a summary of their relationship
column_relationship <- function(df, col_a, col_b) {
  list(
    # cor() calculates the correlation between two numeric vectors
    # use = "complete.obs" tells cor() to only use rows where BOTH columns have values
    # correlation ranges from -1 (strong negative) to 1 (strong positive)
    correlation = cor(df[[col_a]], df[[col_b]], use = "complete.obs"),
    # mean of each column separately, ignoring NAs
    mean_a = mean(df[[col_a]], na.rm = T),
    mean_b = mean(df[[col_b]], na.rm = T),
    # can't use mean_a - mean_b here because those are just labels in the list, not variables
    # so we have to repeat the mean() calls to do the subtraction
    mean_difference = mean(df[[col_a]], na.rm = T) - mean(df[[col_b]], na.rm = T),
    # complete.cases() returns a logical vector — TRUE for each row where BOTH columns are non-NA
    # sum() counts the TRUEs to give us the number of complete rows
    n_complete = sum(complete.cases(df[[col_a]], df[[col_b]]))
  )
}

column_relationship(mtcars, "mpg", "hp")
```

---

## Challenge 7: Multi-column analysis report

Recall your `analyze_column(df, col, threshold = NULL)` function from Challenge 5 of the original practice. Now write `analyze_columns(df, cols, threshold = NULL)` that:

1. Takes a data frame and a **vector** of column names (`cols`).
2. Calls `analyze_column` on each column.
3. Returns a named list where each element is the analysis result for one column.

Test it on `airquality` with `cols = c("Ozone", "Solar.R", "Wind", "Temp")`. Then extract just the `pct_above_threshold` for `"Ozone"` to confirm the structure works.

> **Tip:** This is a wrapper that applies an existing function repeatedly. You'll need a `for` loop and a results list that grows by name (like in Challenge 2 of the original practice).

```{r ORIGINAL7}
# analyze_column <- function(df, col, threshold = NULL) {
#   total <- df[[col]]
#   missing <- sum(is.na(df[[col]]))
#   if (is.null(threshold)){
#     threshold <- mean(df[[col]], na.rm = TRUE)
#   }
#   above <- !is.na(total) & total > threshold
#   count <- sum(above)
#   list(
#     column_name = col,
#     n_missing = missing, 
#     threshold_used = threshold,
#     n_above_threshold = count,
#     pct_above_threshold = count / (length(total) - missing)
#   )
# }
# analyze_columns <- function(df, cols, threshold = NULL){
#   list()
#   for (cols in colnames(df)) {
#     paste0(df[[col]])
#     analyze_column(df[[col]])
#   }
# }

```
```{r Corrected7}
#### CORRECTED

analyze_column <- function(df, col, threshold = NULL) {
  total <- df[[col]]
  missing <- sum(is.na(df[[col]]))
  if (is.null(threshold)){
    threshold <- mean(df[[col]], na.rm = TRUE)
  }
  above <- !is.na(total) & total > threshold
  count <- sum(above)
  list(
    column_name = col,
    n_missing = missing, 
    threshold_used = threshold,
    n_above_threshold = count,
    pct_above_threshold = count / (length(total) - missing)
  )
}

analyze_columns <- function(df, cols, threshold = NULL){
  # start with an empty list to store results — one entry per column
  results <- list()
  # loop over cols (the specific columns passed in, NOT all columns in df)
  # col takes on each column name in turn: "Ozone", "Solar.R", etc.
  for (col in cols) {
    # call analyze_column on each column and store the result
    # results[[col]] creates a named entry — e.g. results[["Ozone"]]
    # pass threshold through so analyze_column can use it (or default to NULL)
    results[[col]] <- analyze_column(df, col, threshold)
  }
  # return the full named list — each element is itself a list from analyze_column
  # access nested results with $ chaining: results$Ozone$pct_above_threshold
  results
}

analyze_columns(airquality, c("Ozone", "Solar.R", "Wind", "Temp"))  
analyze_columns(airquality, c("Ozone", "Solar.R", "Wind", "Temp"))$Ozone$pct_above_threshold

```

---

## Challenge 8: Formatted summary printer

Write `print_column_report(df, col)` that prints a formatted summary of a column to the console using `cat()`. The output should look something like:

```
Column: Ozone
  Mean:   42.13
  Median: 31.50
  Min:    1.00
  Max:    168.00
  Missing values: 37
```

Use your `vec_summary()` function internally to compute the stats, then use `cat()` with `"\n"` for line breaks to format the output. `round()` or `sprintf()` can help control the number of decimal places.

Test it on `airquality` with `"Ozone"` and on `mtcars` with `"mpg"`.

> **Tip:** This function is for side effects (printing), not for returning data. You don't need to return anything meaningful — the printed output *is* the point. But you could also have it `invisible()` return the summary list, so the caller can capture it if they want.

```{r original8}
#### ORIGINAL
# vec_summary <- function(x) {
#   list(
#     mean = mean(x, na.rm = TRUE),
#     median = median(x, na.rm = TRUE),
#     min = min(x, na.rm = TRUE),
#     max = max(x, na.rm = TRUE),
#     n_missing = sum(is.na(x))
#   )
# }
# print_column_report <- function(df, col) {
#   cat(col, "\n",
#       "Mean" = mean, "\n",
#       "Median" = median, "\n",
#       "Min" = min, "\n",
#       "Max" = max, "\n",
#       "Missing values" = n_missing)
#   vec_summary(df[[col]])
# }
```
```{r corrected8}
#### CORRECTED
vec_summary <- function(x) {
  list(
    mean = mean(x, na.rm = TRUE),
    median = median(x, na.rm = TRUE),
    min = min(x, na.rm = TRUE),
    max = max(x, na.rm = TRUE),
    n_missing = sum(is.na(x))
  )
}

print_column_report <- function(df, col) {
  # call vec_summary first and store the result so we can pull stats from it
  stats <- vec_summary(df[[col]])
  # cat() prints to the console — unlike print(), it doesn't add quotes or [1]
  # "\n" adds a line break after each line
  cat("Column", col, "\n")
  # stats$mean accesses the "mean" element from the list that vec_summary returned
  # round(x, 2) limits to 2 decimal places for clean display
  cat(" Mean", round(stats$mean, 2), "\n")
  cat(" Median", round(stats$median, 2), "\n")
  cat(" Min", round(stats$min, 2), "\n")
  cat(" Max", round(stats$max, 2), "\n")
  # n_missing is already a whole number so no need to round
  cat(" Missing values", stats$n_missing, "\n")
  # invisible() returns stats silently — the caller won't see it printed
  # but if they do result <- print_column_report(...), they can access it
  # result$mean will give the full unrounded value (42.12931 not 42.13)
  invisible(stats)
}

print_column_report(airquality, "Ozone")           # just prints
result <- print_column_report(airquality, "Ozone")  # prints AND saves
result$mean                                          # access saved stats

```

---

## Challenge 9: A miniature data cleaning pipeline

Put everything together. Write `clean_and_summarize(df, col, threshold = NULL, drop_outliers = FALSE)` that:

1. Pulls the column from the data frame.
2. Counts the original number of values and the number of NAs.
3. If `drop_outliers` is `TRUE`, removes values that are more than 2 standard deviations from the mean (NAs already removed). Otherwise, keeps all non-NA values.
4. If `threshold` is `NULL`, sets it to the mean of the (cleaned) values.
5. Returns a named list with:
   - `column_name`
   - `n_original` (length of the original column)
   - `n_after_cleaning` (length after removing NAs and possibly outliers)
   - `n_dropped` (the difference)
   - `threshold_used`
   - `n_above_threshold`
   - `summary_stats` — a nested list with `mean`, `median`, `min`, `max` of the cleaned values

Test it on `airquality$Ozone` with default settings, then again with `drop_outliers = TRUE`. The mean should change when outliers are dropped.

> **Tip:** To identify outliers more than 2 SDs from the mean, compute `m <- mean(x, na.rm = TRUE)` and `s <- sd(x, na.rm = TRUE)`, then keep values where `abs(x - m) <= 2 * s`. Remember the `is.na()` check comes first.

```{r original9}
##### ORIGINAL
# clean_and_summarize <- function(df, col, threshold = NULL, drop_outliers = F) {
#   df[[col]] # pulls the colum from df
#   original <- sum(df[[col]]) # counts original number of values
#   missing <- sum(is.na(df[[col]])) # counts the number of NA values
#   m <- mean(df[[col]], na.rm = T) # to identify outliers more than 2 sd from mean
#   s <- sd(df[[col]], na.rm = T)
#   if (drop_outliers == T) {
#     keep <- (!(is.na(abs(df[[col]], na.rm = T)))) # remove outliers and save
#   } else {                       # if drop_outliers = F
#     keep <- (!(is.na(df[[col]])))  # save cleaned values
#   } 
#   if (is.null(threshold)) { # if threshold = NULL then set threshold to the 
#     threshold <- mean(keep, na.rm = T)     # mean of the cleaned values
#   }
#   sum_keep = sum(keep) # number of cleaned values
#   above <- keep > threshold # for use to sum number of values above threshold
#   list(
#     column_name = col, # column name
#     n_original = original, # original number of values
#     n_after_cleaning = sum_keep, # number of cleaned values
#     n_dropped = original - sum_keep, # number of values dropped (outliers and NAs)
#     threshold_used = threshold, 
#     n_above_threshold = sum(above), 
#     summary_stats = list(
#       mean = mean(keep),
#       median = median(keep),
#       min = min(keep),
#       max = max(keep)
#     )
#   )
# }
```

```{r corrected9}
#### CORRECTED
# takes a data frame, column name, optional threshold, and outlier toggle
# cleans the column and returns a detailed summary
clean_and_summarize <- function(df, col, threshold = NULL, drop_outliers = FALSE) {
  # WRONG: df[[col]] on its own line does nothing — must store it in a variable
  # RIGHT: save the column as x so we can reference it throughout the function
  x <- df[[col]]
  # WRONG: sum(df[[col]]) adds up all the values (e.g. 5+8+12...)
  # RIGHT: length() counts how many elements are in the vector
  original <- length(x)
  # this was correct — sum(is.na()) counts the NAs
  missing <- sum(is.na(x))
  # these were correct — compute mean and sd for outlier detection
  # calculated on the full column (ignoring NAs) before any cleaning
  m <- mean(x, na.rm = TRUE)
  s <- sd(x, na.rm = TRUE)
  # WRONG: if (drop_outliers = T) — single = is assignment, not comparison
  # RIGHT: if (drop_outliers) works because it's already TRUE or FALSE
  if (drop_outliers) {
    # WRONG: is.na(abs(df[[col]], na.rm = T)) — na.rm doesn't go in abs()
    #        and is.na() gives TRUE/FALSE, not the actual values
    # RIGHT: two conditions combined with &
    #   !is.na(x) — keep non-missing values
    #   abs(x - m) <= 2 * s — keep values within 2 standard deviations of the mean
    #   abs() gives the distance from the mean regardless of direction
    #   this returns the actual VALUES that pass both tests, not TRUE/FALSE
    clean <- x[!is.na(x) & abs(x - m) <= 2 * s]
  } else {
    # WRONG: !is.na(df[[col]]) gives a logical vector (TRUE/FALSE)
    # RIGHT: x[!is.na(x)] gives the actual non-NA values
    clean <- x[!is.na(x)]
  }
  # this logic was correct — if no threshold given, default to mean of cleaned values
  if (is.null(threshold)) {
    # WRONG: mean(keep, na.rm = T) — keep was a logical vector, not values
    # RIGHT: mean(clean) — clean holds the actual values, no na.rm needed
    #        because NAs were already removed above
    threshold <- mean(clean)
  }
  # WRONG: sum(keep) counts TRUEs in a logical vector, not the number of values
  # RIGHT: length(clean) counts how many values survived cleaning
  n_clean <- length(clean)
  # WRONG: keep > threshold compared TRUE/FALSE against a number
  # RIGHT: clean > threshold compares actual values against threshold
  #        sum() counts how many are above
  above <- sum(clean > threshold)
  list(
    column_name = col,
    # WRONG: original was sum() of values, not a count
    # RIGHT: now it's length(), so this works correctly
    n_original = original,
    n_after_cleaning = n_clean,
    # difference tells us how many were removed (NAs + any outliers)
    n_dropped = original - n_clean,
    threshold_used = threshold,
    n_above_threshold = above,
    # nested list for summary statistics
    summary_stats = list(
      # WRONG: mean(keep) was averaging TRUE/FALSE values
      # RIGHT: mean(clean) averages the actual cleaned numbers
      # no na.rm needed — NAs were already stripped out
      mean = mean(clean),
      median = median(clean),
      min = min(clean),
      max = max(clean)
    )
  )
}

clean_and_summarize(airquality, "Ozone")
clean_and_summarize(airquality, "Ozone", drop_outliers = TRUE)
```

---

## Challenge 10: Bonus — functions that take functions

This is a step toward how `lapply()` and friends work (which is the next chapter). Write `summarize_with(x, func, ...)` that takes a vector `x`, a function `func`, and any extra arguments via `...`. It should call `func(x, ...)` and return the result.

So `summarize_with(airquality$Ozone, mean, na.rm = TRUE)` would call `mean(airquality$Ozone, na.rm = TRUE)`. And `summarize_with(mtcars$mpg, median)` would call `median(mtcars$mpg)`.

This might feel pointless — why wrap a function call inside another function call? The power is that *the function itself becomes an argument*, so the same wrapper can compute means, medians, sums, or anything else without you rewriting it.

Test it with at least three different functions: `mean`, `median`, and `sd`.

```{r}
#### got it right
summarize_with <- function(x, func, ...) {
  result <- func(x, ...)
  result
}

summarize_with(airquality$Ozone, mean, na.rm = TRUE)
summarize_with(mtcars$mpg, median)
summarize_with(mtcars$mpg, sd)
```
