---
title: "R Programming Practice (Part 3)"
author: "Tim Irwin"
date: "2026-04-17"
output:
  pdf_document: default
  html_document: default
---

# R Programming Practice (Part 3)

This practice set is based on *R Programming for Data Science* by Roger D. Peng (pages 44–64):

1. Dates and Times
2. Control Structures (if/else, for, while, repeat, next, break)
3. Functions

Unlike the last practice set, this one is built around **scenarios** using R's **built-in datasets**. Instead of telling you exactly what code to write, each exercise gives you a question to answer or a problem to solve. That means there's usually more than one right approach — what matters is that your answer works and the logic is sound.

Write your answers inside the R chunks below.

You may want to start with:

```{r}
rm(list = ls())
```

## A quick reference for built-in datasets used in this practice

All of these come pre-loaded with R — no installation needed. Run `?datasetname` in the console to see documentation. You can also type the dataset name on its own line to view it.

- `airquality` — Daily air quality measurements in New York, May–September 1973.
- `mtcars` — Fuel consumption and 10 aspects of design/performance for 32 cars.
- `iris` — Measurements of 150 iris flowers across 3 species.
- `women` — Average heights and weights for American women aged 30–39.
- `ChickWeight` — Weight of chicks on different diets over time.
- `Nile` — Annual flow of the river Nile, 1871–1970.

---

# 1. Dates and Times

## Scenario: You're analyzing log data

### Exercise 1.1

Your colleague gives you three character strings representing dates: `"2024-01-15"`, `"2024-06-30"`, and `"2025-03-22"`. Convert them into proper `Date` objects so you can do date arithmetic on them.

```{r}
# Your code here

```

### Exercise 1.2

Using the dates you created above, figure out how many days passed between the earliest and latest dates. (Hint: subtraction works on dates.)

```{r}
# Your code here

```

### Exercise 1.3

Get today's date using `Sys.Date()`. Then find out what day of the week it was 100 days ago. (Hint: look up `weekdays()`.)

```{r}
# Your code here

```

### Exercise 1.4

You're given the following timestamps as character strings representing when a server received requests:

```
"2024-11-05 08:23:41"
"2024-11-05 14:17:03"
"2024-11-05 22:05:19"
```

Convert them to `POSIXct` objects and calculate the time difference (in hours) between the first and last request.

```{r}
# Your code here

```

### Exercise 1.5

Look at the `airquality` dataset. It has `Month` and `Day` columns but no proper date column. Create a new column called `Date` in a copy of the dataset that combines these into real `Date` objects (assume the year is 1973). Hint: you can build a date string with `paste()` and then use `as.Date()`.

```{r}
# Your code here

```

---

# 2. Control Structures

## Scenario: Classifying and filtering data

### Exercise 2.1

Write an `if/else` block that takes a single number `x` and prints:

- `"positive"` if x is greater than 0
- `"negative"` if x is less than 0
- `"zero"` if x is exactly 0

Test it with the values 5, -3, and 0. (You can either change `x` each time or wrap your logic in a function later.)

```{r}
# Your code here

```

### Exercise 2.2

The `mtcars` dataset has a column `mpg` (miles per gallon). Write a `for` loop that iterates through each row and prints a message like `"The Mazda RX4 gets 21 mpg"`. Hint: you can get the car names with `rownames(mtcars)`.

```{r}
# Your code here

```

### Exercise 2.3

Using `mtcars` again, write a `for` loop that only prints the car name for cars that have 6 cylinders. (Check the `cyl` column.) This combines a `for` loop with `if`.

```{r}
# Your code here

```

### Exercise 2.4

Use a `while` loop to find the first year in the `Nile` dataset where the river's annual flow drops below 700. The dataset is a time series starting in 1871, so you'll need to track both the value and the year. Print out the year and the flow value when you find it. Hint: you can convert a time series to a vector with `as.numeric(Nile)` and get the years with `time(Nile)`.

```{r}
# Your code here

```

### Exercise 2.5

Write a `for` loop that iterates through the numbers 1 to 20, but uses `next` to skip multiples of 3, and uses `break` to stop the loop entirely once it hits a number greater than 15. Print each number that doesn't get skipped.

```{r}
# Your code here

```

### Exercise 2.6

The `iris` dataset has 150 rows. Write a `for` loop that goes through each row and classifies the flower based on its `Petal.Length`:

- If Petal.Length < 2: print `"small"`
- If Petal.Length is between 2 and 5: print `"medium"`
- If Petal.Length >= 5: print `"large"`

To avoid flooding your output with 150 lines, only print the classification for the first 10 rows.

```{r}
# Your code here

```

---

# 3. Functions

## Scenario: Building reusable tools for data analysis

### Exercise 3.1

Write a function called `fahrenheit_to_celsius` that takes a temperature in Fahrenheit and returns the equivalent in Celsius. Formula: `C = (F - 32) * 5/9`. Test it with 32, 100, and 212.

```{r}
# Your code here

```

### Exercise 3.2

Write a function called `describe_number` that takes a single number and returns a character string like `"5 is positive and odd"` or `"-4 is negative and even"`. Test it with several values. Hint: the modulo operator `%%` returns the remainder after division.

```{r}
# Your code here

```

### Exercise 3.3

Write a function called `summarize_vector` that takes a numeric vector and returns a named list containing the mean, median, min, max, and number of NAs. Make sure it handles NAs correctly (i.e., the mean/median should exclude them). Test it on `airquality$Ozone`.

```{r}
# Your code here

```

### Exercise 3.4

Write a function called `car_report` that takes the `mtcars` dataset and a car name (as a character string), and prints out a sentence describing that car, like:

`"The Mazda RX4 has 6 cylinders, 110 horsepower, and gets 21 mpg."`

Test it with at least two different car names.

```{r}
# Your code here

```

### Exercise 3.5

Write a function called `above_average` that takes a data frame and the name of a numeric column (as a character string), and returns the subset of rows where that column's value is above the column's mean. Test it on `mtcars` with the column `"mpg"` — you should get back all the cars that are more fuel-efficient than average.

Hint: you can access a column by name with `df[[colname]]`.

```{r}
# Your code here

```

### Exercise 3.6

Write a function with a **default argument**. The function `greet` should take a `name` argument and a `greeting` argument that defaults to `"Hello"`. It should return a string like `"Hello, Tim!"`. Test it both with and without specifying the greeting.

```{r}
# Your code here

```

### Exercise 3.7

Write a function called `filter_species` that takes the `iris` dataset and a species name (e.g., `"setosa"`), and returns only the rows for that species. Then use it to count how many rows there are for each of the three species. (You don't need a loop — just call the function three times.)

```{r}
# Your code here

```

---

# Mini Challenges

## Challenge 1: Temperature trend

Using the `airquality` dataset, write a loop that goes through each of the 5 months (May = 5 through September = 9) and prints the average temperature (`Temp` column) for that month. The output should look something like:

```
Month 5: average temp = 65.5
Month 6: average temp = 79.1
...
```

Remember to handle missing values.

```{r}
# Your code here

```

## Challenge 2: Your first real function

Write a function called `bmi_category` that takes a height (in inches) and weight (in pounds) and returns a BMI category. The BMI formula (using imperial units) is:

`BMI = (weight / height^2) * 703`

Categories:

- BMI < 18.5: `"underweight"`
- 18.5 ≤ BMI < 25: `"normal"`
- 25 ≤ BMI < 30: `"overweight"`
- BMI ≥ 30: `"obese"`

Then use `mapply()` or a `for` loop to apply your function to every row of the `women` dataset and see how many rows fall into each category.

```{r}
# Your code here

```

## Challenge 3: Count NAs across a data frame

Write a function called `count_nas` that takes a data frame and returns a named vector showing how many NAs are in each column. Test it on `airquality`. The result should look like:

```
 Ozone Solar.R    Wind    Temp   Month     Day 
    37       7       0       0       0       0 
```

Hint: think about using `sapply()` — it applies a function to each element (in this case, each column) of a list or data frame.

```{r}
# Your code here

```

## Challenge 4: Chick weight growth

The `ChickWeight` dataset records chick weights over time under different diets. Write code to answer: **what was the average final weight of chicks on each of the 4 diets?**

The "final weight" is the weight at the last time point for each chick (Time = 21 for most, though some chicks died earlier). Use functions, loops, or any combination you like.

```{r}
# Your code here

```

## Challenge 5: A real scenario

You're given a vector of customer ages: `ages <- c(22, 45, 17, 33, 65, 28, 71, 19, 50, 39, NA, 55)`.

Write a single function called `age_summary` that takes such a vector and returns a list with:

- the count of adults (18+)
- the count of minors (under 18)
- the count of seniors (65+)
- the count of missing values
- the average age (excluding NAs)

Test it on the `ages` vector above.

```{r}
# Your code here

```
