---
title: "R Programming Practice (Part 5): Functions"
author: "Tim Irwin"
date: "2026-04-27"
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
---

# R Programming Practice (Part 5): Functions

This practice set covers functions from *R Programming for Data Science* by Roger D. Peng (pages 56–64):

1. Basic function syntax
2. Default arguments and argument matching
3. Functions that operate on vectors
4. Functions with control flow (combining if/else and loops from the last practice set)
5. Functions that operate on data frames
6. Advanced topics: lazy evaluation, `NULL` defaults, the `...` argument

Each section starts with a **lesson** that teaches the concept with a worked example, followed by **exercises** that apply what you just learned. The exercises **build on each other** — each one extends or combines the previous ones.

You may want to start with:

```{r}
rm(list = ls())
```

> **Tip:** Build functions incrementally. Get the simplest version working first, then add features one at a time. If a function isn't working, test the body of it line by line in the console with concrete values before wrapping it back up.

## Datasets used in this practice

- `airquality` — Daily air quality measurements in New York, May–September 1973.
- `mtcars` — Fuel consumption and 10 aspects of design/performance for 32 cars.
- `iris` — Measurements of 150 iris flowers across 3 species.

---

# 1. Basic Functions

## Lesson 1.1: Function syntax

In R, functions are created with the `function()` directive and stored as objects, just like vectors or data frames. The general structure is:

```
my_function <- function(arg1, arg2) {
  # body — code that does something
}
```

Here's a complete, working example. This function takes no arguments and just prints a message to the console:

```{r}
welcome <- function() {
  print("Welcome to R!")
}

welcome()
```

Things to notice:

- We define the function with `function() { ... }` and assign it to `welcome`.
- We **call** the function with `welcome()` — the empty parentheses are required even when there are no arguments.
- The curly braces `{}` enclose the body. R will run everything inside them when the function is called.
- Functions are themselves R objects. You can confirm this with `class(welcome)` — it returns `"function"`.

### Exercise 1.1

Write a function called `say_hello` that takes no arguments and prints `"Hello, world!"`. Call it.

```{r}
say_hello <- function() {
  print("Hello, world!")
}
say_hello()
```

---

## Lesson 1.2: Adding arguments

Functions become useful when they take **arguments** — inputs the function works with. Each argument acts like a placeholder variable available inside the function body.

Here's `welcome` extended to accept a name:

```{r}
welcome_user <- function(user) {
  message <- paste0("Welcome, ", user, "!")
  print(message)
}

welcome_user("Roger")
welcome_user("Maya")

```

`paste0()` glues strings together with no separator between them. `paste()` does the same but adds a space by default — both are useful, depending on what you want.

### Exercise 1.2

Modify your `say_hello` function from Exercise 1.1 to take a `name` argument and print `"Hello, <name>!"` (substituting the actual name). Test it with your own name.

```{r}
say_hello <- function(name) {
  message <- paste0("Hello ", name, "!")
  print(message)
}
say_hello("Tim")

```

---

## Lesson 1.3: Return vs print

This is one of the most important concepts about functions. **Printing** shows something on the screen but doesn't make it available for use later. **Returning** gives a value back to whatever called the function so it can be assigned, passed along, or used in another calculation.

In R, the return value of a function is always the value of the **last expression** in the body. You typically don't need to write `return()` explicitly.

Compare these two functions:

```{r}
prints_only <- function(n) {
  print(n * 3)
}

returns_value <- function(n) {
  n * 3
}

a <- prints_only(7)   # 21 prints, but `a` doesn't get the value
a                     # NULL — nothing useful was returned

b <- returns_value(7) # nothing prints, but `b` gets 21
b
```

> **Tip:** When in doubt, **return** values rather than print them. Returned values can be assigned, passed to other functions, or used in arithmetic. Printed values can only be looked at.

### Exercise 1.3

Modify your function from Exercise 1.2 again so that instead of printing the greeting, it **returns** it as a string. Store the result in a variable called `msg` and then print `msg`.

```{r}
say_hello <- function(name) {
  message <- paste0("Hello ", name, "!")
}
msg <- say_hello("Tim")
msg
```

---

## Lesson 1.4: Multiple arguments

Functions can take any number of arguments, separated by commas. Here's one that takes two:

```{r}
area_of_rectangle <- function(length, width) {
  length * width
}

area_of_rectangle(4, 5)
area_of_rectangle(10, 3)
```

The body uses both arguments, and the result of `length * width` is the last expression — so it gets returned automatically.

### Exercise 1.4

Write a function called `double_it` that takes a number `x` and returns `x` multiplied by 2. Test it with `double_it(5)` and `double_it(-3)`.

```{r}
double_it <- function(x) {
  x * 2
}
double_it(5)
double_it(-3)

```

### Exercise 1.5

Generalize Exercise 1.4: write `multiply_by(x, factor)` that takes two numbers and returns their product. Test it with `multiply_by(5, 3)`.

```{r}
multiply_by <- function(x, factor) {
  x * factor
}

multiply_by(5, 3)
```

---

# 2. Default Arguments and Argument Matching

## Lesson 2.1: Default values

You can give arguments **default values** in the function definition. If the caller doesn't supply a value for that argument, R uses the default. Defaults are set with `=` inside the `function()` parentheses:

```{r}
celsius_to_fahrenheit <- function(c, precision = 1) {
  result <- (c * 9/5) + 32
  round(result, precision)
}

celsius_to_fahrenheit(20)              # uses default precision of 1 → 68
celsius_to_fahrenheit(20, precision = 3)   # overrides default → 68.000
celsius_to_fahrenheit(36.5)             # body temperature → 97.7
```

Defaults are useful when an argument is *usually* the same value, but you occasionally want to override it. This relieves the user from having to specify every argument every time.

### Exercise 2.1

Take your `multiply_by` function from Exercise 1.5 and add a default value of `2` for `factor`. Now `multiply_by(5)` should return 10, but `multiply_by(5, 3)` should still return 15.

```{r}
multiply_by <- function(x, factor = 2) {
  x * factor
}

multiply_by(5)

multiply_by(5, 3)
```

---

## Lesson 2.2: How R matches arguments

R is flexible about how you supply arguments. Given a function `f(x, y)`, you can call it:

- **Positionally:** `f(1, 2)` — first value goes to `x`, second to `y`
- **By name:** `f(x = 1, y = 2)` — explicit assignment
- **By name in any order:** `f(y = 2, x = 1)` — order doesn't matter when using names
- **Mixed:** `f(1, y = 2)` — positional first, then named

```{r}
power <- function(base, exponent) {
  base ^ exponent
}

power(2, 10)                              # positional
power(base = 2, exponent = 10)            # named
power(exponent = 10, base = 2)            # named, reverse order
power(2, exponent = 10)                   # mixed
```

All four return the same answer (1024).

> **Tip:** Named arguments make code more readable, especially when a function has many arguments or when the name clarifies what a value represents. For functions with just one or two arguments, positional is fine.

### Exercise 2.2

Call your `multiply_by` function from Exercise 2.1 four different ways and verify they all give the same answer:

1. Positional: `multiply_by(5, 3)`
2. Named: `multiply_by(x = 5, factor = 3)`
3. Named, reverse order: `multiply_by(factor = 3, x = 5)`
4. Mixed: `multiply_by(5, factor = 3)`

```{r}
multiply_by(5, 3)
multiply_by(x = 5, factor = 3)
multiply_by(factor = 3, x = 5)
multiply_by(5, factor = 3)
```

---

## Lesson 2.3: Required vs optional arguments

Arguments without a default are **required** — the function will throw an error if they're missing. Arguments with defaults are **optional**. R's convention is to put required arguments first and optional ones (with defaults) after.

```{r}
order_total <- function(price, quantity, tax_rate = 0.08) {
  subtotal <- price * quantity
  subtotal + (subtotal * tax_rate)
}

order_total(20, 3)                       # uses default tax_rate
order_total(20, 3, tax_rate = 0.10)      # overrides default
```

If you call `order_total(20)` without supplying `quantity`, R will error with `"argument 'quantity' is missing, with no default"`.

### Exercise 2.3

Write a function called `discount_price(price, discount_pct = 10)` that returns the price after applying a percentage discount. So `discount_price(100)` should return 90, and `discount_price(100, 25)` should return 75.

```{r}
discount_price <- function(price, discount_pct = 10) {
  price - discount_pct
}
discount_price(100)
discount_price(100, 25)

```

---

# 3. Functions That Operate on Vectors

## Lesson 3.1: Functions and vectors

Most useful R functions operate on whole vectors, not single values. When you write `mean(x)`, `x` is a whole vector. Your custom functions should usually be designed the same way.

Here's a function that uses a `for` loop to compute the **product** of all elements in a vector:

```{r}
my_product <- function(x) {
  result <- 1                       # start at 1, not 0 (multiplying by 0 zeros everything)
  for (i in seq_along(x)) {
    result <- result * x[i]
  }
  result
}

my_product(c(2, 3, 4))   # 2 * 3 * 4 = 24
my_product(c(5, 10))     # 50
```

The pattern: initialize an accumulator (`result <- 1`), loop over the vector, update the accumulator each iteration, and the final value is the result. This is the same loop pattern you used in the control structures practice — just wrapped in a function so you can reuse it.

### Exercise 3.1

Write a function called `vec_sum(x)` that takes a numeric vector and returns the sum of its elements. **Don't use `sum()`** — write it with a `for` loop. Test it with `c(1, 2, 3, 4, 5)` (should return 15).

```{r}
vec_sum <- function(x) {
  result <- 0
  for (i in seq_along(x)) {
    result <- result + x[i]
  }
  result
}

vec_sum(c(1, 2, 3, 4, 5))
```

---

## Lesson 3.2: Adding NA handling with arguments

Real data has missing values. A common pattern is to add an `na.rm` argument that controls whether NAs are skipped. R's built-in functions like `mean()` and `sum()` follow this convention.

```{r}
my_max <- function(x, na.rm = FALSE) {
  best <- -Inf                     # start at negative infinity
  for (i in seq_along(x)) {
    if (is.na(x[i])) {
      if (na.rm) {
        next                       # skip this value
      } else {
        return(NA)                 # if any NA and na.rm=FALSE, result is NA
      }
    }
    if (x[i] > best) {
      best <- x[i]
    }
  }
  best
}

my_max(c(3, 7, NA, 4))                # NA (default behavior)
my_max(c(3, 7, NA, 4), na.rm = TRUE)  # 7
```

The `if (is.na(x[i]))` check catches NAs *before* trying to use them in arithmetic, which would otherwise produce errors or propagate NAs.

### Exercise 3.2

Modify `vec_sum` from Exercise 3.1 to handle NAs. Add an argument `na.rm` that defaults to `FALSE`. If `TRUE`, skip NA values. If `FALSE` and there are NAs, return `NA`. Test it with `c(1, 2, NA, 4, 5)` both ways.

```{r}
vec_sum <- function(x, na.rm = FALSE) {
  result <- 0
  for (i in seq_along(x)) {
    if (is.na(x[i])) {
      if (na.rm) {
        next
      } else {
        return(NA)
      }
    }
    result <- result + x[i]
  }
  result
}

vec_sum(c(1, 2, NA, 4, 5))
vec_sum(c(1, 2, NA, 4, 5), na.rm = TRUE)
```

### Exercise 3.3

Test your `vec_sum` from Exercise 3.2 on `airquality$Ozone` with `na.rm = TRUE`. Compare your result to `sum(airquality$Ozone, na.rm = TRUE)` — they should match.

```{r}
vec_sum(airquality$Ozone, na.rm = TRUE)
sum(airquality$Ozone, na.rm = TRUE)

```

---

## Lesson 3.3: Returning multiple values with a named list

What if your function needs to return more than one thing? You can't return multiple separate values, but you *can* wrap them in a named list:

```{r}
spread_info <- function(x) {
  list(
    smallest = min(x, na.rm = TRUE),
    largest = max(x, na.rm = TRUE),
    range_size = max(x, na.rm = TRUE) - min(x, na.rm = TRUE)
  )
}

result <- spread_info(c(15, 22, 8, 31, 14))
result               # see the whole list
result$smallest      # extract a specific value
result$range_size
```

The names act as labels, so the caller knows what each value represents and can extract specific ones with `$`. This is the standard R idiom for returning multiple values from one function.

### Exercise 3.4

Write a function called `vec_summary(x)` that takes a numeric vector and returns a named list with five elements: `mean`, `median`, `min`, `max`, and `n_missing` (the count of NAs). Make sure the mean/median/min/max calculations exclude NAs.

```{r}
vec_summary <- function(x) {
  list(
    mean = mean(x, na.rm = TRUE),
    median = median(x, na.rm = TRUE),
    min = min(x, na.rm = TRUE),
    max = max(x, na.rm = TRUE),
    n_missing = sum(is.na(x))
  )
}

```

### Exercise 3.5

Apply `vec_summary` from Exercise 3.4 to two different columns: `airquality$Wind` and `airquality$Ozone`. Then extract just the `mean` from each result using `$`.

```{r}
vec_summary(airquality$Wind)
vec_summary(airquality$Ozone)
vec_summary(airquality$Wind)$mean
vec_summary(airquality$Ozone)$mean

```

---

# 4. Functions with Control Flow

This section combines functions with the if/else and loops you practiced in Part 4.

## Lesson 4.1: if/else inside functions

Putting `if/else` inside a function lets you make decisions based on the input. The pattern is the same as standalone if/else — it's just inside the function body:

```{r}
weather_advice <- function(temp) {
  if (temp > 85) {
    "Stay hydrated and seek shade"
  } else if (temp < 40) {
    "Bundle up — it's cold"
  } else {
    "Comfortable weather"
  }
}

weather_advice(92)
weather_advice(35)
weather_advice(70)
```

Notice there's no explicit `return()` — whichever branch executes, its value is the last expression evaluated, so that's what gets returned.

### Exercise 4.1

Write a function called `classify(x)` that takes a single number and returns:

- `"positive"` if x > 0
- `"negative"` if x < 0
- `"zero"` if x == 0

Test it with several values.

```{r}
classify <- function(x) {
  if (x > 0) {
    "positive"
  } else if (x < 0) {
    "negative"
  } else {
    "zero"
  }
}
classify(9)
classify(-12)
classify(0)
```

---

## Lesson 4.2: Handling NA inside functions

You hit this in your control structures practice: comparisons with `NA` produce `NA`, which can break `if` statements. The fix is to **check for NA first** before doing any other comparison:

```{r}
rate_speed <- function(mph) {
  if (is.na(mph)) {
    return("unknown")        # early return — done
  }
  
  if (mph > 70) {
    "fast"
  } else if (mph < 30) {
    "slow"
  } else {
    "moderate"
  }
}

rate_speed(80)
rate_speed(25)
rate_speed(NA)
```

This uses an explicit `return()` for the NA case so the function exits immediately. After that, the rest of the function can safely assume `mph` is a real number.

### Exercise 4.2

Modify `classify` from Exercise 4.1 to also handle `NA`. If `x` is missing, return `"missing"`. Check `is.na()` *first*, before the other comparisons.

```{r}
classify <- function(x) {
  if (is.na(x)){
    return("missing")
  }
  
  if (x > 0) {
    "positive"
  } else if (x < 0) {
    "negative"
  } else {
    "zero"
  }
}

classify(9)
classify(-12)
classify(0)
classify(NA)
```

---

## Lesson 4.3: Vectorizing — making it work on whole vectors

The functions above take a single number. To apply that logic to a whole vector, you can write a second function that loops over the vector and calls the single-value function on each element:

```{r}
rate_speed <- function(mph) {
  if (is.na(mph)) return("unknown")
  if (mph > 70) "fast"
  else if (mph < 30) "slow"
  else "moderate"
}

# Vectorized version
rate_speed_all <- function(mphs) {
  results <- character(length(mphs))      # pre-allocate result vector
  for (i in seq_along(mphs)) {
    results[i] <- rate_speed(mphs[i])     # call inner function
  }
  results
}

rate_speed_all(c(80, 25, 55, NA, 10))
```

Two patterns to notice:

1. **Pre-allocate** the result vector with `character(length(mphs))`. This is more efficient than growing a vector inside a loop.
2. **Calling your own function from inside another function** is encouraged — it keeps each function small and focused on one job.

### Exercise 4.3

Write `classify_all(x)` that takes a **vector** of numbers and returns a **vector** of classifications. Use a `for` loop internally, and call your `classify` function from Exercise 4.2 inside the loop.

Test it with `c(5, -3, 0, NA, 10)` — you should get back `c("positive", "negative", "zero", "missing", "positive")`.

```{r}
# Your code here
classify <- function(x) {
  if (is.na(x)) return("missing")
  if (x > 0) "positive"
  else if (x < 0) "negative"
  else "zero"
}

classify_all <- function(x) {
  results <- character(length(x))
  for (i in seq_along(x)) {
    results[i] <- classify(x[i])
  }
  results
}

(classify_all(c(5, -3, 0, NA, 10)))
```

### Exercise 4.4

Apply `classify_all` from Exercise 4.3 to `airquality$Temp - 75` (each day's temperature relative to 75°F). Use `table()` on the result to see how many days were above, at, or below 75.

```{r}
table(classify_all(airquality$Temp - 75))
```

### Exercise 4.5

Write a function called `categorize_mpg(x)` that takes a numeric vector of mpg values and returns a character vector classifying each:

- `"efficient"` if mpg > 25
- `"average"` if 15 ≤ mpg ≤ 25
- `"thirsty"` if mpg < 15

Apply it to `mtcars$mpg` and use `table()` to see the counts.

```{r}
mpg <- function(x) {
  if (is.na(x)) return("missing")
  if (x > 25) "efficient"
  else if (x < 15) "thirsty"
  else "average"
}

categorize_mpg <- function(x) {
  results <- character(length(x)) 
  for (i in seq_along(x)) {
    results[i] <- mpg(x[i])
  }
  results
}

categorize_mpg(mtcars$mpg)
table(categorize_mpg(mtcars$mpg))
```

---

# 5. Functions That Operate on Data Frames

## Lesson 5.1: Accessing a column by a variable

When you have the column name in a variable (rather than typing it literally), you need to use `[[ ]]`, not `$`:

```{r}
get_column <- function(df, col) {
  df[[col]]    # works — uses the value of `col`
  # df$col     # WRONG — looks for a literal column called "col"
}

head(get_column(mtcars, "mpg"))
head(get_column(iris, "Species"))
```

The `$` operator only works with literal names — it doesn't evaluate variables. This is one of the most common gotchas when writing functions that work with data frames.

### Exercise 5.1

Write a function called `column_mean(df, col)` where `df` is a data frame and `col` is a column name as a character string. The function should return the mean of that column with NAs removed. Test it on `mtcars` with `"mpg"`, and on `airquality` with `"Ozone"`.

```{r}
column_mean <- function(df, col) {
  mean(df[[col]], na.rm = TRUE)
}
column_mean(mtcars, "mpg")
column_mean(airquality, "Ozone")
```

---

## Lesson 5.2: Building functions out of other functions

Once you have a small, focused function, you can use it as a building block for larger ones. Here's a function that uses our earlier `spread_info` to summarize a column:

```{r}
spread_info <- function(x) {
  list(
    smallest = min(x, na.rm = TRUE),
    largest = max(x, na.rm = TRUE),
    range_size = max(x, na.rm = TRUE) - min(x, na.rm = TRUE)
  )
}

column_spread <- function(df, col) {
  spread_info(df[[col]])
}

column_spread(mtcars, "mpg")
column_spread(airquality, "Wind")
```

The new function pulls the column out of the data frame and hands it off to the existing function. This is **composition** — building bigger tools from smaller ones. It keeps each function small enough to reason about and reuse.

### Exercise 5.2

Write `column_summary(df, col)` that uses your `vec_summary` function from Exercise 3.4 internally — pull the column out of the data frame, then pass it to `vec_summary`. Test it on `mtcars` with `"hp"`.

```{r}
vec_summary <- function(x) {
  list(
    mean = mean(x, na.rm = TRUE),
    median = median(x, na.rm = TRUE),
    min = min(x, na.rm = TRUE),
    max = max(x, na.rm = TRUE),
    n_missing = sum(is.na(x))
  )
}

column_summary <- function(df, col) {
  vec_summary(df[[col]])
}

column_summary(mtcars, "hp")
```

---

## Lesson 5.3: Filtering rows based on a condition

You can write functions that subset a data frame based on values in a column:

```{r}
top_quartile <- function(df, col) {
  cutoff <- quantile(df[[col]], 0.75, na.rm = TRUE)
  df[df[[col]] >= cutoff, ]
}

# Cars in the top quartile of horsepower
top_quartile(mtcars, "hp")
```

Note the subsetting syntax: `df[df[[col]] >= cutoff, ]` keeps rows where the condition is true (the `,` with nothing after it means "all columns").

### Exercise 5.3

Write a function called `above_average(df, col)` that returns the subset of rows where the value in column `col` is above the mean of that column (NAs removed). Test it on `mtcars` with `"mpg"` — you should get back the more fuel-efficient cars.

```{r}
above_average <- function(df, col) {
  average <- mean(df[[col]], na.rm = TRUE)
  df[df[[col]] > average, ]
}

above_average(mtcars, "mpg")
```

---

## Lesson 5.4: NULL defaults for "compute later"

Sometimes a default value depends on the input data and can't be computed when the function is defined (because the data doesn't exist yet). The trick is to use `NULL` as a placeholder default and check for it inside the function with `is.null()`:

```{r}
filter_above <- function(df, col, threshold = NULL) {
  if (is.null(threshold)) {
    threshold <- mean(df[[col]], na.rm = TRUE)   # default to the column's mean
  }
  df[df[[col]] > threshold, ]
}

# Use the default (column mean):
result1 <- filter_above(mtcars, "mpg")
nrow(result1)

# Override with a specific threshold:
result2 <- filter_above(mtcars, "mpg", threshold = 25)
nrow(result2)
```

The `is.null()` check tells you whether the caller supplied a value or not. This is a common pattern in R — you'll see it in many built-in functions.

### Exercise 5.4

Write `above_threshold(df, col, threshold = NULL)` that returns the subset of rows where `col` exceeds `threshold`. If `threshold` is `NULL`, default to the column's mean.

Test it both ways: with an explicit threshold and with the default.

```{r}
above_treshold <- function(df, col, threshold = NULL) {
  if(is.null(threshold)) {
    threshold <- mean(df[[col]], na.rm = TRUE)
  }
  df[df[[col]] > threshold, ]
}

result3 <- filter_above(mtcars, "hp")
nrow(result3)

result4 <- filter_above(mtcars, "hp", threshold = 180)
nrow(result4)

```

---

# 6. Advanced Topics

## Lesson 6.1: Lazy evaluation

R only evaluates function arguments **when they're actually used** in the body. This means you can sometimes call a function without supplying all the arguments — as long as the missing one is never referenced:

```{r}
quirky <- function(p, q) {
  p * 10           # only `p` is used; `q` is never touched
}

quirky(7)          # works, even though we didn't supply q
```

If we tried this on a function that uses `q`, R would error when it hit the line that references `q`. The key idea: R doesn't check whether arguments exist up front — it only checks when each one is actually needed.

> **Tip:** Lazy evaluation is more of a quirk than a feature to rely on. Don't write functions that depend on it — always supply all required arguments. But it's good to know about, because it explains some confusing error messages where part of the function output appears before the error.

### Exercise 6.1

Write a function `lazy_demo(a, b)` that prints `a` but never references `b`. Then call it with just one argument: `lazy_demo(5)`. It should work without error.

```{r}
lazy_demo <- function(a, b) {
  a
}
lazy_demo(5)

```

---

## Lesson 6.2: The `...` argument

The `...` (three dots) argument is a special placeholder that lets your function accept additional arguments and pass them along to another function. This is useful when you're building a wrapper around an existing function and don't want to copy its entire argument list.

```{r}
verbose_max <- function(x, ...) {
  cat("Finding the max...\n")
  result <- max(x, ...)              # pass `...` through to max()
  cat("Done!\n")
  result
}

# Now we can pass na.rm = TRUE through to max(),
# even though our function doesn't declare na.rm:
verbose_max(airquality$Ozone, na.rm = TRUE)
```

The caller can use any argument that `max()` accepts (`na.rm`, etc.), and `...` will forward them on. You'll see `...` used in many built-in R functions, often as the very first argument to functions like `paste()` and `cat()` that can accept any number of inputs.

### Exercise 6.2

Write `my_mean(x, ...)` that calls `mean(x, ...)` internally. Test it with `my_mean(airquality$Ozone, na.rm = TRUE)`.

```{r}
my_mean <- function(x, ...) {
  mean(x, ...)
}
my_mean(airquality$Ozone, na.rm = TRUE)
```

### Exercise 6.3

Write a function `describe_column(df, col, ...)` that prints the column name and then calls `summary(df[[col]], ...)`. The `...` allows callers to pass additional arguments to `summary()`. Test it on `mtcars` with `"mpg"`.

```{r}
describe_column <- function(df, col, ...) {
  cat(col, "\n")
  summary(df[[col]], ...)
}

describe_column(mtcars, "mpg")
```

---

# Mini Challenges

These tie everything together. Each one combines multiple concepts from above.

## Challenge 1: BMI calculator with units

Write `bmi_calc(weight, height, units = "imperial")` that calculates BMI:

- If `units = "imperial"`: weight in pounds, height in inches. Formula: `(weight / height^2) * 703`
- If `units = "metric"`: weight in kg, height in meters. Formula: `weight / height^2`

The function should return a **named list** with two elements: `bmi` (the numeric value) and `category` (one of `"underweight"`, `"normal"`, `"overweight"`, `"obese"` based on standard cutoffs: <18.5, 18.5–25, 25–30, ≥30).

If `units` is anything other than `"imperial"` or `"metric"`, halt with `stop("Invalid units")`.

Test it with both unit systems.

> **Tip:** `stop("message")` halts execution and reports an error. It's how you signal that something is wrong with the inputs, instead of returning a wrong answer silently.

```{r, error = TRUE}
bmi_calc <- function(weight, height, units = "imperial") {
  if (units == "imperial") {
    bmi <- (weight/height^2) * 703
  } else if (units == "metric"){
    bmi <- weight/height^2
  } else {
    stop("Invalid units")
  }
  
  category <- if (bmi < 18.5) {
    "underweight"
  } else if (bmi >= 18.5 & bmi < 25) {
    "normal"
  } else if (bmi >= 25 & bmi < 30) {
    "overweight"
  } else {
    "obese"
  } 
list(
  bmi = bmi,
  category = category
)
}

bmi_calc(150, 67)                        # imperial, should be ~23.5, normal
bmi_calc(68, 1.73, units = "metric")     # metric, should be ~22.7, normal
bmi_calc(100, 70)                        # imperial, underweight
bmi_calc(300, 65)                        # imperial, obese
bmi_calc(150, 67, units = "banana")      # should trigger your stop() error
```

---

## Challenge 2: Apply to multiple columns

Using your `vec_summary` function from Exercise 3.4, write a `for` loop that applies it to **every column** of `mtcars` and stores the results in a named list called `all_summaries`. Then access `all_summaries$mpg$mean` to verify it worked.

Hint: `colnames(mtcars)` gives you all the column names to loop over.

```{r}
vec_summary <- function(x) {
  list(
    mean = mean(x, na.rm = TRUE),
    median = median(x, na.rm = TRUE),
    min = min(x, na.rm = TRUE),
    max = max(x, na.rm = TRUE),
    n_missing = sum(is.na(x))
  )
}

all_summaries <- list()
for (col in colnames(mtcars)) {
   all_summaries[[col]] <-  vec_summary(mtcars[[col]])
}
all_summaries$mpg$mean
```

---

## Challenge 3: Smart summary

Write a function called `smart_summary(x)` that adapts its behavior to the input type:

- If `x` is numeric, return `vec_summary(x)` (your function from Exercise 3.4)
- If `x` is character or factor, return a `table()` of the values
- If `x` is logical, return the count of `TRUE` and `FALSE` values

Use `if/else if` and the `is.numeric()`, `is.character()`, `is.factor()`, `is.logical()` test functions.

Test it on `iris$Sepal.Length` (numeric), `iris$Species` (factor), and a logical vector you make yourself.

```{r}
vec_summary <- function(x) {
  list(
    mean = mean(x, na.rm = TRUE),
    median = median(x, na.rm = TRUE),
    min = min(x, na.rm = TRUE),
    max = max(x, na.rm = TRUE),
    n_missing = sum(is.na(x))
  )
}

smart_summary <- function(x) {
  if (is.numeric(x)) {
    vec_summary(x)
  } else if (is.character(x) | is.factor(x)) {
    table(x)
  } else if (is.logical(x)) {
    table(x)
  }
}


smart_summary(iris$Sepal.Length)
smart_summary(iris$Species)
smart_summary(c(TRUE, FALSE, TRUE, TRUE, FALSE))

```

---

## Challenge 4: Refactor

Here's some working code that's repetitive — the same calculation done twice on different columns:

```r
ozone_above_50 <- airquality$Ozone[!is.na(airquality$Ozone) & airquality$Ozone > 50]
ozone_above_50_count <- length(ozone_above_50)
ozone_above_50_mean <- mean(ozone_above_50)

wind_above_15 <- airquality$Wind[!is.na(airquality$Wind) & airquality$Wind > 15]
wind_above_15_count <- length(wind_above_15)
wind_above_15_mean <- mean(wind_above_15)
```

Refactor it into a single function `above_threshold_stats(x, threshold)` that takes a vector and threshold and returns a named list with `count` and `mean` of values above the threshold. Then reproduce the same results in just two function calls.

> **Tip:** Refactoring repetitive code into functions is one of the highest-value skills in programming. Any time you find yourself copying code with small changes, that's a sign a function should exist.

```{r}
above_threshold_stats <- function(x, threshold) {
  above <- x[!is.na(x) & x > threshold]
  list(
    mean = mean(above, na.rm = TRUE),
    count = sum(!is.na(above))
  )
}

above_threshold_stats(airquality$Ozone, 50)
above_threshold_stats(airquality$Wind, 15)

```

---

## Challenge 5: A complete pipeline

Combine everything. Write `analyze_column(df, col, threshold = NULL)` that:

1. Pulls the column from the data frame.
2. Counts how many NAs are in it.
3. If `threshold` is `NULL`, sets it to the column's mean (NAs removed).
4. Counts how many non-missing values are above the threshold.
5. Returns a named list with: `column_name`, `n_missing`, `threshold_used`, `n_above_threshold`, and `pct_above_threshold` (as a percentage of non-missing values).

Test it on `airquality` with `"Ozone"` (let it use the default threshold) and on `mtcars` with `"hp"` and `threshold = 200`.

```{r}
analyze_column <- function(df, col, threshold = NULL) {
  total <- df[[col]]
  missing <- sum(is.na(df[[col]]))
  if (is.null(threshold)){
    threshold <- mean(df[[col]], na.rm = TRUE)
  }
  above <- !is.na(total) & total > threshold
  count <- sum(above)
  list(
    column_name = col,
    n_missing = missing, 
    threshold_used = threshold,
    n_above_threshold = count,
    pct_above_threshold = count / (length(total) - missing)
  )
}

analyze_column(airquality, "Ozone")
analyze_column(mtcars, "hp", threshold = 200)
```
