---
title: "R Programming Practice (Part 5): Functions"
author: "Tim Irwin"
date: "2026-04-27"
output:
  pdf_document: default
  html_document: default
---

# R Programming Practice (Part 5): Functions

This practice set covers functions from *R Programming for Data Science* by Roger D. Peng (pages 56–64), but with a twist: each section starts with a **lesson** that teaches the concept with a worked example, followed by **exercises** that apply what you just learned.

The exercises **build on each other** — each one extends or combines a previous one. If you get stuck, look back at the lesson examples or the earlier exercises in the same section.

You may want to start with:

```{r}
rm(list = ls())
```

> **Tip:** Read the lesson, run the example code in your console to see it work, then attempt the exercise. If you get stuck on an exercise, revisit the lesson example and try modifying it line by line.

## Datasets used in this practice

- `airquality` — Daily air quality measurements in New York, May–September 1973.
- `mtcars` — Fuel consumption and performance for 32 cars.
- `iris` — Measurements of 150 iris flowers across 3 species.

---

# 1. Basic Functions

## Lesson 1.1: Function syntax

A function in R is created with the `function()` keyword and assigned to a name with `<-`. The general structure looks like this:

```
my_function <- function(arg1, arg2) {
  # body of function — code that does something
}
```

Here's a complete, working example. This function takes no arguments and just prints a greeting:

```{r}
say_hi <- function() {
  print("Hi there!")
}

say_hi()  # call the function
```

Things to notice:

- We define the function with `function() { ... }` and assign it to `say_hi`.
- We **call** the function with `say_hi()` — the empty parentheses are required.
- The curly braces `{}` enclose the body of the function.

### Exercise 1.1

Write a function called `say_hello` that takes no arguments and prints `"Hello, world!"`. Call it.

```{r}
# Your code here

```

---

## Lesson 1.2: Adding arguments

Functions become useful when they take **arguments** — inputs the function works with. Here's `say_hi` extended to accept a name:

```{r}
greet_person <- function(name) {
  message <- paste0("Hi there, ", name, "!")
  print(message)
}

greet_person("Tim")
greet_person("Roger")
```

`paste0()` glues strings together with no separator between them. `paste()` does the same but adds a space by default — both are useful.

### Exercise 1.2

Modify your `say_hello` function from Exercise 1.1 to take a `name` argument and print `"Hello, <name>!"` (substituting the actual name). Test it with your own name.

```{r}
# Your code here

```

---

## Lesson 1.3: Return vs print

This is one of the most important concepts about functions. **Printing** shows something on the screen, but doesn't make it available for use later. **Returning** gives a value back to whatever called the function.

Compare these two functions:

```{r}
f_print <- function(x) {
  print(x * 2)   # prints to screen
}

f_return <- function(x) {
  x * 2          # returns the value
}

a <- f_print(5)   # 10 prints, but `a` doesn't get the value
a                 # NULL — nothing useful was returned

b <- f_return(5)  # nothing prints, but `b` gets 10
b
```

R automatically returns the value of the **last expression** in the function body. You don't need to write `return()` explicitly — that's only needed when you want to exit a function early.

> **Tip:** When in doubt, **return** values rather than print them. Returned values can be assigned, passed to other functions, or used in calculations. Printed values can only be looked at.

### Exercise 1.3

Modify your function from Exercise 1.2 again so that instead of printing the greeting, it **returns** it as a string. Store the result in a variable called `msg` and then print `msg`.

```{r}
# Your code here

```

---

## Lesson 1.4: Multiple arguments

Functions can take any number of arguments, separated by commas:

```{r}
add_three <- function(a, b, c) {
  a + b + c
}

add_three(1, 2, 3)
add_three(10, 20, 30)
```

### Exercise 1.4

Write a function called `double_it` that takes a number `x` and returns `x` multiplied by 2. Test it with `double_it(5)` and `double_it(-3)`.

```{r}
# Your code here

```

### Exercise 1.5

Generalize Exercise 1.4: write `multiply_by(x, factor)` that takes two numbers and returns their product. Test it with `multiply_by(5, 3)`.

```{r}
# Your code here

```

---

# 2. Default Arguments and Argument Matching

## Lesson 2.1: Default values

You can give arguments **default values** in the function definition. If the caller doesn't supply that argument, R uses the default. Defaults are set with `=` inside the `function()` parentheses:

```{r}
power_of <- function(base, exponent = 2) {
  base ^ exponent
}

power_of(5)        # uses default exponent of 2 → 25
power_of(5, 3)     # overrides default → 125
```

Defaults are useful when an argument is *usually* the same value, but you occasionally want to override it.

### Exercise 2.1

Take your `multiply_by` function from Exercise 1.5 and add a default value of `2` for `factor`. Now `multiply_by(5)` should return 10, but `multiply_by(5, 3)` should still return 15.

```{r}
# Your code here

```

---

## Lesson 2.2: How R matches arguments

R is flexible about how you supply arguments. Given a function `f(x, y)`, you can call it:

- **Positionally:** `f(1, 2)` — first value goes to `x`, second to `y`
- **By name:** `f(x = 1, y = 2)` — explicit assignment
- **By name in any order:** `f(y = 2, x = 1)` — order doesn't matter when using names
- **Mixed:** `f(1, y = 2)` — positional first, then named

```{r}
divide <- function(numerator, denominator) {
  numerator / denominator
}

divide(10, 2)                              # positional
divide(numerator = 10, denominator = 2)    # named
divide(denominator = 2, numerator = 10)    # named, reverse order
divide(10, denominator = 2)                # mixed
```

All four return the same answer.

> **Tip:** Named arguments make code more readable, especially when a function has many arguments or when the name clarifies what a value represents. For functions with just one or two arguments, positional is fine.

### Exercise 2.2

Call your `multiply_by` function from Exercise 2.1 four different ways and verify they all give the same answer:

1. Positional: `multiply_by(5, 3)`
2. Named: `multiply_by(x = 5, factor = 3)`
3. Named, reverse order: `multiply_by(factor = 3, x = 5)`
4. Mixed: `multiply_by(5, factor = 3)`

```{r}
# Your code here

```

---

## Lesson 2.3: Required vs optional arguments

Arguments without a default are **required** — the function will throw an error if they're missing. Arguments with defaults are **optional**.

```{r}
greet <- function(name, greeting = "Hello") {
  paste0(greeting, ", ", name, "!")
}

greet("Tim")                       # uses default greeting
greet("Tim", greeting = "Howdy")   # overrides default
```

By convention, required arguments come first in the function definition, and optional ones (with defaults) come after.

### Exercise 2.3

Write a function called `discount_price(price, discount_pct = 10)` that returns the price after applying a percentage discount. So `discount_price(100)` should return 90, and `discount_price(100, 25)` should return 75.

```{r}
# Your code here

```

---

# 3. Functions That Operate on Vectors

## Lesson 3.1: Functions and vectors

Most useful R functions operate on whole vectors, not single values. When you write `mean(x)`, `x` is a whole vector. Your custom functions should usually be designed the same way.

Here's a function that uses a `for` loop to sum the elements of a vector:

```{r}
my_total <- function(x) {
  total <- 0
  for (i in seq_along(x)) {
    total <- total + x[i]
  }
  total
}

my_total(c(10, 20, 30, 40))
```

The pattern: initialize an accumulator (`total <- 0`), loop over the vector, update the accumulator each iteration, and the final value is the result. This is the same loop pattern from your control structures practice — just wrapped in a function.

### Exercise 3.1

Write a function called `vec_sum(x)` that takes a numeric vector and returns the sum of its elements. **Don't use `sum()`** — write it with a `for` loop. Test it with `c(1, 2, 3, 4, 5)` (should return 15).

```{r}
# Your code here

```

---

## Lesson 3.2: Adding NA handling with arguments

Real data has missing values. A common pattern is to add an `na.rm` argument that controls whether NAs are skipped. R's built-in functions like `mean()` and `sum()` follow this convention.

```{r}
my_count <- function(x, na.rm = FALSE) {
  count <- 0
  for (i in seq_along(x)) {
    if (is.na(x[i])) {
      if (na.rm) {
        next   # skip this value
      } else {
        return(NA)  # if any NA and na.rm=FALSE, result is NA
      }
    }
    count <- count + 1
  }
  count
}

my_count(c(1, 2, NA, 4))                 # NA (default behavior)
my_count(c(1, 2, NA, 4), na.rm = TRUE)   # 3
```

The `if (is.na(x[i]))` check catches NAs *before* trying to use them in arithmetic, which would otherwise propagate the NA.

### Exercise 3.2

Modify `vec_sum` from Exercise 3.1 to handle NAs. Add an argument `na.rm` that defaults to `FALSE`. If `TRUE`, skip NA values. If `FALSE` and there are NAs, return `NA`. Test it with `c(1, 2, NA, 4, 5)` both ways.

```{r}
# Your code here

```

### Exercise 3.3

Test your `vec_sum` from Exercise 3.2 on `airquality$Ozone` with `na.rm = TRUE`. Compare your result to `sum(airquality$Ozone, na.rm = TRUE)` — they should match.

```{r}
# Your code here

```

---

## Lesson 3.2: Returning multiple values with a named list

What if your function needs to return more than one thing? You can't return multiple separate values, but you *can* wrap them in a named list:

```{r}
range_info <- function(x) {
  list(
    minimum = min(x, na.rm = TRUE),
    maximum = max(x, na.rm = TRUE),
    spread = max(x, na.rm = TRUE) - min(x, na.rm = TRUE)
  )
}

result <- range_info(c(3, 7, 1, 9, 4))
result               # see the whole list
result$minimum       # extract a specific value
result$spread
```

The names act as labels, so the caller knows what each value represents and can extract specific ones with `$`.

### Exercise 3.4

Write a function called `vec_summary(x)` that takes a numeric vector and returns a named list with five elements: `mean`, `median`, `min`, `max`, and `n_missing` (the count of NAs). Make sure the mean/median/min/max calculations exclude NAs.

```{r}
# Your code here

```

### Exercise 3.5

Apply `vec_summary` from Exercise 3.4 to two different columns: `airquality$Wind` and `airquality$Ozone`. Then extract just the `mean` from each result using `$`.

```{r}
# Your code here

```

---

# 4. Functions with Control Flow

This section combines functions with the if/else and loops you practiced in Part 4.

## Lesson 4.1: if/else inside functions

Putting `if/else` inside a function lets you make decisions based on the input. The pattern is the same as standalone if/else — it's just inside the function body:

```{r}
sign_of <- function(x) {
  if (x > 0) {
    "positive"
  } else if (x < 0) {
    "negative"
  } else {
    "zero"
  }
}

sign_of(5)
sign_of(-3)
sign_of(0)
```

Notice there's no explicit `return()` — whichever branch executes, its value is the last expression evaluated, so that's what gets returned.

### Exercise 4.1

Write a function called `classify(x)` that takes a single number and returns:

- `"positive"` if x > 0
- `"negative"` if x < 0
- `"zero"` if x == 0

Test it with several values.

```{r}
# Your code here

```

---

## Lesson 4.2: Handling NA inside functions

You hit this in your control structures practice: comparisons with `NA` produce `NA`, which can break `if` statements. The fix is to **check for NA first** before doing any other comparison:

```{r}
describe_temp <- function(temp) {
  if (is.na(temp)) {
    return("missing")    # early return — done
  }
  
  if (temp > 80) {
    "hot"
  } else if (temp < 60) {
    "cold"
  } else {
    "moderate"
  }
}

describe_temp(85)
describe_temp(55)
describe_temp(NA)
```

This uses an explicit `return()` for the NA case so the function exits immediately. After that, the rest of the function can safely assume `temp` is a real number.

### Exercise 4.2

Modify `classify` from Exercise 4.1 to also handle `NA`. If `x` is missing, return `"missing"`. Check `is.na()` *first*, before the other comparisons.

```{r}
# Your code here

```

---

## Lesson 4.3: Vectorizing — make it work on whole vectors

The `classify` function takes a single number. To apply it to a whole vector, you can write a second function that loops over the vector and calls `classify` on each element:

```{r}
sign_of <- function(x) {
  if (x > 0) "positive"
  else if (x < 0) "negative"
  else "zero"
}

# Vectorized version
sign_of_all <- function(x) {
  results <- character(length(x))    # pre-allocate result vector
  for (i in seq_along(x)) {
    results[i] <- sign_of(x[i])      # call inner function
  }
  results
}

sign_of_all(c(5, -3, 0, 10, -7))
```

Two patterns to notice:

1. **Pre-allocate** the result vector with `character(length(x))`. This is more efficient than growing a vector inside a loop.
2. **Calling your own function from inside another function** is encouraged — it keeps each function small and focused on one job.

### Exercise 4.3

Write `classify_all(x)` that takes a **vector** of numbers and returns a **vector** of classifications. Use a `for` loop internally, and call your `classify` function from Exercise 4.2 inside the loop.

Test it with `c(5, -3, 0, NA, 10)` — you should get back `c("positive", "negative", "zero", "missing", "positive")`.

```{r}
# Your code here

```

### Exercise 4.4

Apply `classify_all` from Exercise 4.3 to `airquality$Temp - 75` (each day's temperature relative to 75°F). Use `table()` on the result to see how many days were above, at, or below 75.

```{r}
# Your code here

```

### Exercise 4.5

Write a function called `categorize_mpg(x)` that takes a numeric vector of mpg values and returns a character vector classifying each:

- `"efficient"` if mpg > 25
- `"average"` if 15 ≤ mpg ≤ 25
- `"thirsty"` if mpg < 15

Apply it to `mtcars$mpg` and use `table()` to see the counts.

```{r}
# Your code here

```

---

# 5. Functions That Operate on Data Frames

## Lesson 5.1: Accessing a column by a variable

When you have the column name in a variable (rather than typing it literally), you need to use `[[ ]]`, not `$`:

```{r}
get_column <- function(df, col) {
  df[[col]]    # works — uses the value of `col`
  # df$col     # WRONG — looks for a literal column called "col"
}

head(get_column(mtcars, "mpg"))
```

The `$` operator only works with literal names — it doesn't evaluate variables. This is one of the most common gotchas when writing functions that work with data frames.

### Exercise 5.1

Write a function called `column_mean(df, col)` where `df` is a data frame and `col` is a column name as a character string. The function should return the mean of that column with NAs removed. Test it on `mtcars` with `"mpg"`, and on `airquality` with `"Ozone"`.

```{r}
# Your code here

```

---

## Lesson 5.2: Building functions out of other functions

Once you have a small, focused function, you can use it as a building block for larger ones. Here's a function that uses our earlier `range_info` to summarize a column:

```{r}
range_info <- function(x) {
  list(
    minimum = min(x, na.rm = TRUE),
    maximum = max(x, na.rm = TRUE),
    spread = max(x, na.rm = TRUE) - min(x, na.rm = TRUE)
  )
}

column_range <- function(df, col) {
  range_info(df[[col]])
}

column_range(mtcars, "mpg")
```

The new function pulls the column out of the data frame and hands it off to the existing function. This is composition — building bigger tools from smaller ones.

### Exercise 5.2

Write `column_summary(df, col)` that uses your `vec_summary` function from Exercise 3.4 internally — pull the column out of the data frame, then pass it to `vec_summary`. Test it on `mtcars` with `"hp"`.

```{r}
# Your code here

```

---

## Lesson 5.3: Filtering rows based on a condition

You can write functions that subset a data frame based on values in a column:

```{r}
top_quartile <- function(df, col) {
  cutoff <- quantile(df[[col]], 0.75, na.rm = TRUE)
  df[df[[col]] >= cutoff, ]
}

# Cars in the top quartile of horsepower
top_quartile(mtcars, "hp")
```

Note the subsetting syntax: `df[df[[col]] >= cutoff, ]` keeps rows where the condition is true (the `,` with nothing after it means "all columns").

### Exercise 5.3

Write a function called `above_average(df, col)` that returns the subset of rows where the value in column `col` is above the mean of that column (NAs removed). Test it on `mtcars` with `"mpg"` — you should get back the more fuel-efficient cars.

```{r}
# Your code here

```

---

## Lesson 5.4: NULL defaults for "compute later"

Sometimes a default value depends on the input data and can't be computed when the function is defined. The trick is to use `NULL` as a placeholder default and check for it inside the function:

```{r}
filter_above <- function(df, col, threshold = NULL) {
  if (is.null(threshold)) {
    threshold <- mean(df[[col]], na.rm = TRUE)   # default to the mean
  }
  df[df[[col]] > threshold, ]
}

# Use the default (column mean):
result1 <- filter_above(mtcars, "mpg")
nrow(result1)

# Override with a specific threshold:
result2 <- filter_above(mtcars, "mpg", threshold = 25)
nrow(result2)
```

The `is.null()` check tells you whether the caller supplied a value or not.

### Exercise 5.4

Write `above_threshold(df, col, threshold = NULL)` that returns the subset of rows where `col` exceeds `threshold`. If `threshold` is `NULL`, default to the column's mean.

Test it both ways: with an explicit threshold and with the default.

```{r}
# Your code here

```

---

# 6. Advanced Topics

## Lesson 6.1: Lazy evaluation

R only evaluates function arguments when they're actually used in the body. This means you can sometimes call a function without supplying all the arguments — as long as the missing one is never referenced:

```{r}
demo <- function(a, b) {
  a * 2          # only `a` is used; `b` is never touched
}

demo(5)          # works, even though we didn't supply b
```

If we tried `demo(5)` on a function that uses `b`, R would error when it hit the line that references `b`.

> **Tip:** Lazy evaluation is more of a quirk than a feature to rely on. Don't write functions that depend on it — always supply all required arguments. But it's good to know about, because it explains some confusing error messages.

### Exercise 6.1

Write a function `lazy_demo(a, b)` that prints `a` but never references `b`. Then call it with just one argument: `lazy_demo(5)`. It should work without error.

```{r}
# Your code here

```

---

## Lesson 6.2: The `...` argument

The `...` (three dots) argument is a special placeholder that lets your function accept additional arguments and pass them along to another function. This is useful when you're building a wrapper:

```{r}
verbose_mean <- function(x, ...) {
  cat("Calculating mean...\n")
  result <- mean(x, ...)              # pass `...` through to mean()
  cat("Done!\n")
  result
}

# Now we can pass na.rm = TRUE through to mean(),
# even though our function doesn't declare na.rm:
verbose_mean(airquality$Ozone, na.rm = TRUE)
```

The caller can use any argument that `mean()` accepts (`na.rm`, `trim`, etc.), and `...` will forward them on.

### Exercise 6.2

Write `my_mean(x, ...)` that calls `mean(x, ...)` internally. Test it with `my_mean(airquality$Ozone, na.rm = TRUE)`.

```{r}
# Your code here

```

### Exercise 6.3

Write a function `describe_column(df, col, ...)` that prints the column name and then calls `summary(df[[col]], ...)`. The `...` allows callers to pass additional arguments to `summary()`. Test it on `mtcars` with `"mpg"`.

```{r}
# Your code here

```

---

# Mini Challenges

These tie everything together. Each one combines multiple concepts from above.

## Challenge 1: BMI calculator with units

Write `bmi_calc(weight, height, units = "imperial")` that calculates BMI:

- If `units = "imperial"`: weight in pounds, height in inches. Formula: `(weight / height^2) * 703`
- If `units = "metric"`: weight in kg, height in meters. Formula: `weight / height^2`

The function should return a **named list** with two elements: `bmi` (the numeric value) and `category` (one of `"underweight"`, `"normal"`, `"overweight"`, `"obese"` based on standard cutoffs: <18.5, 18.5–25, 25–30, ≥30).

If `units` is anything other than `"imperial"` or `"metric"`, halt with `stop("Invalid units")`.

Test it with both unit systems.

> **Tip:** `stop("message")` halts execution and reports an error. It's how you signal that something is wrong with the inputs, instead of returning a wrong answer silently.

```{r}
# Your code here

```

---

## Challenge 2: Apply to multiple columns

Using your `vec_summary` function from Exercise 3.4, write a `for` loop that applies it to **every column** of `mtcars` and stores the results in a named list called `all_summaries`. Then access `all_summaries$mpg$mean` to verify it worked.

Hint: `colnames(mtcars)` gives you all the column names to loop over.

```{r}
# Your code here

```

---

## Challenge 3: Smart summary

Write a function called `smart_summary(x)` that adapts its behavior to the input type:

- If `x` is numeric, return `vec_summary(x)` (your function from Exercise 3.4)
- If `x` is character or factor, return a `table()` of the values
- If `x` is logical, return the count of `TRUE` and `FALSE` values

Use `if/else if` and the `is.numeric()`, `is.character()`, `is.factor()`, `is.logical()` test functions.

Test it on `iris$Sepal.Length` (numeric), `iris$Species` (factor), and a logical vector you make yourself.

```{r}
# Your code here

```

---

## Challenge 4: Refactor

Here's some working code that's repetitive — the same calculation done twice on different columns:

```r
ozone_above_50 <- airquality$Ozone[!is.na(airquality$Ozone) & airquality$Ozone > 50]
ozone_above_50_count <- length(ozone_above_50)
ozone_above_50_mean <- mean(ozone_above_50)

wind_above_15 <- airquality$Wind[!is.na(airquality$Wind) & airquality$Wind > 15]
wind_above_15_count <- length(wind_above_15)
wind_above_15_mean <- mean(wind_above_15)
```

Refactor it into a single function `above_threshold_stats(x, threshold)` that takes a vector and threshold and returns a named list with `count` and `mean` of values above the threshold. Then reproduce the same results in just two function calls.

> **Tip:** Refactoring repetitive code into functions is one of the highest-value skills in programming. Any time you find yourself copying code with small changes, that's a sign a function should exist.

```{r}
# Your code here

```

---

## Challenge 5: A complete pipeline

Combine everything. Write `analyze_column(df, col, threshold = NULL)` that:

1. Pulls the column from the data frame.
2. Counts how many NAs are in it.
3. If `threshold` is `NULL`, sets it to the column's mean (NAs removed).
4. Counts how many non-missing values are above the threshold.
5. Returns a named list with: `column_name`, `n_missing`, `threshold_used`, `n_above_threshold`, and `pct_above_threshold` (as a percentage of non-missing values).

Test it on `airquality` with `"Ozone"` (let it use the default threshold) and on `mtcars` with `"hp"` and `threshold = 200`.

```{r}
# Your code here

```
